---
output: 
  github_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# tern

The `tern` R package contains analysis functions to create tables, listings and graphs (TLGs) as used in clinical trials. We also provide [teal](https://github.roche.com/NEST/teal) modules for outputs in `tern` in the [teal.modules.clinical](https://github.roche.com/NEST/teal.modules.clinical) R package.

* Note that currently **tern may not be used for regulatory submissions**

* However you can use `tern` for exploratory analysis and for quality control


# Installation 

## Stable Version

```{r eval=FALSE}
devtools::install_github(
  repo = "NEST/random.cdisc.data",
  ref = "v0.2.0", 
  host = "https://github.roche.com/api/v3",
  upgrade_dependencies = FALSE
)

devtools::install_github("Roche/rtables", ref = "v0.1.1")

devtools::install_github(
  repo = "NEST/tern",
  ref = "v0.6.0", 
  host = "https://github.roche.com/api/v3",
  upgrade_dependencies = FALSE
)
```

# Development

## Issues

We hope that `tern` is useful for your everyday work. If you would like more
TLGs or you need more configuration options for the existing functions then
please fill out a [`github` issue](https://github.roche.com/NEST/tern/issues)
with your request and the reason/context for your request.

* Please do not submit issues with ideas for useful features for current `tern`
functions if you do not necessarily need them for your project. We are aware
that many of `tern`'s functions would benefit from additional arguments in order
to control the outputs in more detail. However, we would like to avoid argument
explosion and one way to do that is to get more experience with `tern` *in the
field* and then slowly grow the flexibility based on need.

* Please do not paste outputs with sensitive information into the issues. If
possible make a minimal working example with `random.cdisc.data` data.


## Collaboration

We would like `tern` to be a collaborative effort and your contributions are
welcome! However please read the following points before you make any
contributions:

* talk to the respective function authors before you start working on changes
other than bug fixes or documentation examples (this includes argument changes)
 
      - please have a look at the [R
      code](https://github.roche.com/NEST/tern/tree/master/R) of a couple
      of functions in `tern`. If you are new to `R` or do not want to spend the
      effort to create similarly general and robust functions then please
      consider contributing to the
      [osprey](https://github.roche.com/Rpackages/osprey) R package.



# Presentations

* [Stats Software Initiative - R Series. June 18, 2018](https://docs.google.com/presentation/d/1OB7MMt3YKzfMJ-gXcGpcRqM8tjbMZWqeEki164L38i4/edit?usp=sharing)

[ghs]: http://pages.github.roche.com/NEST/tern
