---
title: "Design: Best overall response (RSPT01)"
author: "Francois Collin, Daniel Sabanes Bove"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    theme: journal
    highlight: zenburn

---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "svg"
)
```


Configuration
=============

```{r setup, eval = TRUE}
library(rtables)
library(tern)
library(assertthat)

packageVersion("tern")
packageVersion("rtables")
```

Execution
=========

Estimate proportion {.tabset}
-----------------------------

### Application

```{r estimate_proportion, eval = FALSE}
dta_test <- data.frame(X = c(1, 0, 1, 0))
l <- estimate_proportion(lyt = NULL, vars = "X")

build_table(lyt = l, df = dta_test)
```

```{r s_proportion, eval = FALSE}
s_proportion(c(1, 0, 1, 0))
```

### Functions

- `prop_wilson()`
- `prop_clopper_pearson()`
- `prop_wald()`
- `prop_agresti_coull()`
- `prop_jeffreys()`
- `s_proportion()`
- `estimate_proportion()`
- `d_proportion()`

```{r estimate_proportion.R, engine='bash', comment='', eval = FALSE}
cat ../R/estimate_proportion.R
```


Estimate proportion differences {.tabset}
-----------------------------------------

### Application

```{r estimate_proportion_diff_app, eval = FALSE}
# Wald confidence interval

set.seed(2)
rsp <- sample(c(TRUE, FALSE), replace = TRUE, size = 20)
grp <- c(rep("A", 10), rep("B", 10))
prop_diff_wald(rsp = rsp, grp = grp, conf_level = 0.90, correct = FALSE)

# Anderson-Hauck confidence interval

## "Mid" case: 3/4 respond in group A, 1/2 respond in group B.
rsp <- c(TRUE, FALSE, FALSE, TRUE, TRUE, TRUE)
grp <- factor(c("A", "B", "A", "B", "A", "A"), levels = c("B", "A"))
prop_diff_ha(rsp = rsp, grp = grp, conf_level = 0.90)

## Edge case: Same proportion of response in A and B.
rsp <- c(TRUE, FALSE, TRUE, FALSE)
grp <- factor(c("A", "A", "B", "B"), levels = c("A", "B"))
prop_diff_ha(rsp = rsp, grp = grp, conf_level = 0.6)


# Newcombe confidence interval

set.seed(1)
rsp <- c(
  sample(c(TRUE, FALSE), size = 40, prob = c(3/4, 1/4), replace = TRUE),
  sample(c(TRUE, FALSE), size = 40, prob = c(1/2, 1/2), replace = TRUE)
)
grp <- factor(rep(c("A", "B"), each = 40), levels = c("B", "A"))
table(rsp, grp)
prop_diff_nc(rsp = rsp, grp = grp, conf_level = 0.9)


# Cochran-Mantel-Haenszel confidence interval

set.seed(2)
rsp <- sample(c(TRUE, FALSE), 100, TRUE)
grp <- sample(c("Placebo", "Treatment"), 100, TRUE)
grp <- factor(grp, levels = c("Placebo", "Treatment"))
strata_data <- data.frame(
  "f1" = sample(c("a", "b"), 100, TRUE),
  "f2" = sample(c("x", "y", "z"), 100, TRUE),
  stringsAsFactors = TRUE)

prop_diff_cmh(
  rsp = rsp, grp = grp, strata = interaction(strata_data),
  conf_level = 0.90
)


# Summary

## "Mid" case: 4/4 respond in group A, 1/2 respond in group B.
dta <- data.frame(
  rsp = c(TRUE, FALSE, FALSE, TRUE, TRUE, TRUE),
  grp = factor(c("A", "B", "A", "B", "A", "A"), levels = c("B", "A"))
)

s_proportion_diff(
  df = subset(dta, grp == "A"),
  .var = "rsp",
  .ref_group = subset(dta, grp == "B"),
  .in_ref_col = FALSE,
  conf_level = 0.90,
  method = "ha"
)


l <- split_cols_by(lyt = NULL, var = "grp", ref_group = "B") %>%
  estimate_proportion_diff(
    vars = "rsp",
    conf_level = 0.90,
    method = "ha"
  )

build_table(l, df = dta)

```

### Functions

- `check_diff_prop_ci()`
- `d_proportion_diff()`
- `prop_diff_wald()`
- `prop_diff_ha()`
- `prop_diff_nc()`
- `prop_diff_cmh()`
- `s_proportion_diff()`
- `estimate_proportion_diff()`

```{r prop_diff.R, engine='bash', comment='', eval = FALSE}
cat ../R/prop_diff.R
```


Test proportion difference {.tabset}
------------------------------------

### Application

```{r, eval = FALSE}
dta <- data.frame(
  rsp = sample(c(TRUE, FALSE), 100, TRUE),
  grp = factor(rep(c("A", "B"), each = 50)),
  strat = factor(rep(c("V", "W", "X", "Y", "Z"), each = 20))
)

s_test_proportion_diff(
  df = subset(dta, grp == "A"),
  .var = "rsp",
  .ref_group = subset(dta, grp == "B"),
  .in_ref_col = FALSE,
  variables = list(strata = "strat"),
  method = "cmh"
)

l <- split_cols_by(lyt = NULL, var = "grp", ref_group = "B") %>%
  test_proportion_diff(
    vars = "rsp",
    method = "cmh", variables = list(strata = "strat")
  )

build_table(l, df = dta)

```

### Functions

- `prop_chisq()`
- `prop_cmh()`
- `prop_schouten()`
- `prop_fisher()`
- `s_test_proportion_diff()`
- `d_test_proportion_diff()`
- `test_proportion_diff()`

```{r test_proportion_diff.R, engine='bash', comment='', eval = FALSE}
cat ../R/test_proportion_diff.R
```

Estimate Odds Ratio {.tabset}
-----------------------------

### Application

```{r, eval = FALSE}
dta <- data.frame(
  rsp = sample(c(TRUE, FALSE), 100, TRUE),
  grp = factor(rep(c("A", "B"), each = 50))
)

l <- split_cols_by(lyt = NULL, var = "grp", ref_group = "B") %>%
  estimate_odds_ratio(vars = "rsp")

build_table(l, df = dta)
```

### Functions

- `or_glm()`
- `s_odds_ratio()`
- `estimate_odds_ratio()`

```{r odds_ratio.R, engine='bash', comment='', eval = FALSE}
cat ../R/odds_ratio.R
```

### New Additions for Stratified Odds Ratio estimation

- `or_clogit()`
- modifications for Statistics and Layout creating functions

#### `or_clogit`

```{r or_clogit}
library(survival)

#' OR with Strata by Conditional Likelihood
#'
#' The odds ratio is estimated by conditional likelihood using
#' [survival::clogit()]. This is done for the whole data set including all groups, since 
#' the results are not the same as when doing pairwise comparisons between the groups.
#'
#' @inheritParams argument_convention
#' @param data (`data frame`)\cr
#'   includes the three variables `rsp`, `grp` and `strata`.
#' @importFrom survival strata clogit coxph
#'
or_clogit <- function(data, conf_level) {

  assert_that(
    is_df_with_variables(data, list(rsp = "rsp", grp = "grp", strata = "strata")),
    is.logical(data$rsp),
    is_character_or_factor(data$grp),
    is_character_or_factor(data$strata),
    is_proportion(conf_level)
  )
  
  data$grp <- as_factor_keep_attributes(data$grp)
  data$strata <- as_factor_keep_attributes(data$strata)
  
  # Deviation from convention: `survival::strata` must be simply `strata`.
  formula <- stats::as.formula("rsp ~ grp + strata(strata)")
  model_fit <- survival::clogit(formula = formula, data = data)

  # Create a list with one set of OR estimates and CI per coefficient, i.e.
  # comparison of one group vs. the reference group.
  coef_est <- stats::coef(model_fit)
  ci_est <- stats::confint(model_fit, level = conf_level)
  or_ci <- list()
  for (coef_name in names(coef_est)) {
    grp_name <- gsub("^grp", "", x = coef_name)
    or_ci[[grp_name]] <- setNames(
      object = exp(c(coef_est[coef_name], ci_est[coef_name, , drop = TRUE])),
      nm = c("est", "lcl", "ucl")
    )
  }
  list(or_ci = or_ci)
}
```

Note that OR estimates obtained for a
model fitted on data subset is not equivalent to the estimates based on the
complete dataset:

```{r show_diff_or_clogit}
data <- data.frame(
  rsp = as.logical(c(1, 1, 0, 1, 0, 0, 1, 1, 0, 0, 1, 1, 0, 1, 0, 0, 1, 1, 0, 0)),
  grp =    letters[c(1, 1, 1, 2, 2, 2, 3, 3, 3, 3, 1, 1, 1, 2, 2, 2, 3, 3, 3, 3)],
  strata = LETTERS[c(1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2)]
)

data <- subset(data, grp %in% c("a", "b"))
data$grp <- droplevels(data$grp)
or_clogit(data, conf_level = 0.95)

data <- data.frame(
  rsp = as.logical(c(1, 1, 0, 1, 0, 0, 1, 1, 0, 0, 1, 1, 0, 1, 0, 0, 1, 1, 0, 0)),
  grp =    letters[c(1, 1, 1, 2, 2, 2, 3, 3, 3, 3, 1, 1, 1, 2, 2, 2, 3, 3, 3, 3)],
  strata = LETTERS[c(1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2)]
)

data$grp <- droplevels(data$grp)
or_clogit(data, conf_level = 0.95)
```

#### Modified `s_odds_ratio()`

The statistics function needs to use the whole data set for fitting if a stratified analysis is chosen, and then
select the result which is required for this column.

Now we need a `variables` argument which the user can use to specify the stratification and the arm variable.
Note that we need the arm variable since we need to know which to use as the `grp` input for the estimation. 
This was not necessary with the pairwise comparison in `or_glm` since we could work with the subsets of the data frame, i.e. we implictly knew which group has which data already.

```{r s_odds_ratio_strat}
s_odds_ratio <- function(df,
                         .var,
                         .ref_group,
                         .in_ref_col,
                         .df_row,
                         variables = list(arm = NULL, strata = NULL),
                         conf_level = 0.95) {
  y <- list(or_ci = "")

  if (!.in_ref_col) {
    assert_that(
      is_df_with_variables(df, list(rsp = .var)),
      is_df_with_variables(.ref_group, list(rsp = .var)),
      is_proportion(conf_level)
    )
    
    if (is.null(variables$strata)) {
      data <- data.frame(
        rsp = c(.ref_group[[.var]], df[[.var]]),
        grp = factor(
          rep(c("ref", "Not-ref"), c(nrow(.ref_group), nrow(df))),
          levels = c("ref", "Not-ref")
        )
      )
      y <- or_glm(data, conf_level = conf_level)
    } else {
      assert_that(
        is_df_with_variables(.df_row, c(list(rsp = .var), variables))
      )
      data <- data.frame(
        rsp = .df_row[[.var]],
        grp = .df_row[[variables$arm]],
        strata = .df_row[[variables$strata]]
      )
      y_all <- or_clogit(data, conf_level = conf_level)
      this_grp <- as.character(unique(df[[variables$arm]]))
      assert_that(is.string(this_grp))
      y$or_ci <- y_all$or_ci[[this_grp]]
    }
  }

  y$or_ci <- with_label(
    x = y$or_ci,
    label = paste0("Odds Ratio (", 100 * conf_level, "% CI)")
  )
  y
}
```

Let's try it out:

```{r try_new_s_odds_ratio}
data <- data.frame(
  rsp = as.logical(c(1, 1, 0, 1, 0, 0, 1, 1, 0, 0)),
  grp = letters[c(1, 1, 1, 2, 2, 2, 3, 3, 3, 3)],
  strata = letters[c(1, 1, 2, 2, 1, 2, 1, 1, 2, 2)]
)

result <- s_odds_ratio(
  df = subset(data, grp == "b"),
  variables = list(arm = "grp", strata = "strata"),
  .df_row = data,
  .var = "rsp",
  .ref_group = subset(data, grp == "a"),
  .in_ref_col = FALSE
)
```
So it looks like it is working.

#### Test the layout creating function

It seems like we don't need to modify `estimate_odds_ratio`. Let's try - we just need to copy the functions here so
that we override the Formatted analysis function.

```{r test_estimate_or_stratified}
a_odds_ratio <- make_afun(
  s_odds_ratio,
  .formats = c(or_ci = "xx.xx (xx.xx - xx.xx)"),
  .indent_mods = c(or_ci = 1L)
)

estimate_odds_ratio <- function(lyt,
                                vars,
                                ...,
                                show_labels = "hidden",
                                .stats = NULL,
                                .formats = NULL,
                                .labels = NULL,
                                .indent_mods = NULL) {
  afun <- make_afun(
    a_odds_ratio,
    .stats = .stats,
    .formats = .formats,
    .labels = .labels,
    .indent_mods = .indent_mods
  )

  analyze(
    lyt,
    vars,
    afun = afun,
    extra_args = list(...),
    show_labels = show_labels
  )
}

data <- data.frame(
  rsp = as.logical(c(1, 1, 0, 1, 0, 0, 1, 1, 0, 0)),
  grp = letters[c(1, 1, 1, 2, 2, 2, 3, 3, 3, 3)],
  strata = letters[c(1, 1, 2, 2, 1, 2, 1, 1, 2, 2)]
)

result <- basic_table() %>%
  split_cols_by(var = "grp", ref_group = "a") %>%
  estimate_odds_ratio(vars = "rsp", variables = list(arm = "grp", strata = "strata")) %>%
  build_table(df = data)
result

```

So this seems to work ok.

Estimate multinomial response
-----------------------------

### Application

```{r, eval = FALSE}
dta_test <- data.frame(
  USUBJID = paste0("S", 1:12),
  ARM     = rep(LETTERS[1:3], each = 4),
  AVAL    = c(A = c(1, 1, 1, 1), B = c(0, 0, 1, 1), C = c(0, 0, 0, 0))
)
dta_test$AVALC <- as.factor(c(
  "Complete Response (CR)", "Partial Response (PR)"
)[dta_test$AVAL + 1])

lyt <- split_cols_by(lyt = NULL, var = "ARM") %>%
  estimate_multinomial_response(var = "AVALC")

build_table(lyt = lyt, df = dta_test)
```

### Functions

- `d_onco_rsp_label()`
- `h_prop_ci()`
- `s_multinomial_response()`
- `estimate_multinomial_response()`

```{r estimate_multinomial_rsp.R, engine='bash', comment='', eval = FALSE}
cat ../R/estimate_multinomial_rsp.R
```

Stacked layers
==============

```{r, eval = FALSE}
# Preparation of the test case.
library(dplyr)
library(random.cdisc.data)
adsl <- radsl(cached = TRUE)
adrs <- radrs(cached = TRUE)

# Select a response parameter endpoint.
adrs <- adrs %>% filter(PARAMCD == "INVET")

# Recode multinomial response with standard oncology labels.
levels(adrs$AVALC) <- d_onco_rsp_label(levels(adrs$AVALC))

# Response variable.
adrs <- adrs %>%
  select(ARMCD, AVALC, SEX, STRATA1, STRATA2) %>%
  mutate(
    is_rsp = AVALC %in% c("Complete Response (CR)", "Partial Response (PR)")
  )

l <- split_cols_by(lyt = NULL, var = "ARMCD", ref_group = "ARM A") %>%
  add_colcounts() %>%
  estimate_proportion(vars = "is_rsp") %>%
  estimate_proportion_diff(
    vars = "is_rsp", show_labels = "visible",
    var_labels = "Unstratified Analysis"
  ) %>%
  test_proportion_diff(vars = "is_rsp", method = "schouten") %>%
  estimate_odds_ratio(vars = "is_rsp") %>%
  estimate_multinomial_response(var = "AVALC")

build_table(l, adrs) %>% as_html
```


References
==========
