---
title: "Design: Patient Disposition Table (DST01)"
author: "Daniel Sabanes Bove"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Design of DST01}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "svg"
)
```

## Configuration

```{r setup}
library(rtables)
library(tern)

packageVersion("tern")
packageVersion("rtables")
```

```{r data, results = "asis", eval = TRUE}

# This specific ADSL is currently not part of RCD.
set.seed(1, kind = "Mersenne-Twister")

library(dplyr)
library(random.cdisc.data)
library(assertthat)

#nolint start
adsl0 <- ex_adsl %>%
  mutate(
    COMPSTUD = sample(c('Y','N'), size=nrow(ex_adsl), replace = TRUE) %>% as.factor,
    STUDONS = sample(
      c('Alive: On Treatment', 'Alive: In Follow-up', NA), 
      size = nrow(ex_adsl), 
      replace = TRUE) %>% as.factor,
    STDDRS = sample(c('Death', 'Lost To Follow-Up',
                      'Protocol Violation', 'Withdrawal By Subject',
                      'Other'),
                    size=nrow(ex_adsl),
                    replace = TRUE) %>% as.factor,
    GOTTRT = ifelse(!is.na(ACTARMCD), 'Y', 'N') %>% as.factor,
    DISTRTFL = sample(c('Y','N'),
                      size=nrow(ex_adsl),
                      replace = TRUE) %>% as.factor,
    TRTDRS = sample(c('ADVERSE EVENT', 'PROGRESSIVE DISEASE',
                      'PHYSICIAN DECISION', 'LACK OF EFFICACY',
                      'OTHER'),
                    size=nrow(ex_adsl),
                    replace = TRUE) %>% as.factor,
    STUDONS = case_when(COMPSTUD == 'N' ~ STUDONS),
    STDDRS = case_when(COMPSTUD == 'N' & is.na(STUDONS) ~ STDDRS),
    DISSTDFL = case_when(!is.na(STDDRS) ~ "Y"),
    DISTRTFL = case_when(GOTTRT == 'Y' ~ DISTRTFL),
    TRTDRS = case_when(DISTRTFL == 'Y' ~ TRTDRS),
    DRSCAT = case_when(
      TRTDRS %in% c('ADVERSE EVENT', 'PHYSICIAN DECISION') ~ "Safety",
      !is.na(TRTDRS) ~ "Other"
    )
  ) %>% var_relabel(COMPSTUD = "Complete Study",
                    STUDONS = "On-study Status",
                    DISSTDFL = "Discontinued Study",
                    STDDRS = "Reason for Study \r\nDiscontinuation",
                    GOTTRT = "Received Treatment",
                    DISTRTFL = "Discontinued Treatment",
                    TRTDRS = "Reason for Treatment \r\nDiscontinuation",
                    DRSCAT = "Subcategory for Treatment Discontinuation"
  )
#nolint end
adsl0
```

## Variant 1: Simple Patient Disposition Table

### Objective

We want something like this:

````markdown
                                 A: Drug X             B: Placebo          C: Combination         All Patients 
                                  (N=134)               (N=134)               (N=132)               (N=400)    
---------------------------------------------------------------------------------------------------------------
Completed study                 64 (47.76%)           64 (47.76%)           69 (52.27%)           197 (49.25%) 
Discontinued study              24 (17.91%)           14 (10.45%)           20 (15.15%)            58 (14.5%)  
  Death                          4 (2.99%)             2 (1.49%)             5 (3.79%)             11 (2.75%)  
  Lost To Follow-Up              4 (2.99%)             3 (2.24%)             1 (0.76%)               8 (2%)    
  Other                          3 (2.24%)             4 (2.99%)             4 (3.03%)             11 (2.75%)  
  Protocol Violation             6 (4.48%)             2 (1.49%)             6 (4.55%)             14 (3.5%)   
  Withdrawal By Subject          7 (5.22%)             3 (2.24%)             4 (3.03%)             14 (3.5%)   
````

### Design Proposition

The final functions should make this workable:

```{r target, eval = FALSE}
basic_table() %>%
  split_cols_by("ARM", split_fun = add_overall_level("All Patients", first = FALSE)) %>%
  add_colcounts() %>%
  count_values("COMPSTUD", values = "Y", var_labels = "Completed Study") %>%
  split_rows_by("DISSTDFL", split_fun = keep_split_levels("Y")) %>%
  summarize_row_groups(label_fstr = "Discontinued Study") %>%
  summarize_rows("STDDRS") %>%
  build_table(ADSL0)
```

### Function `count_value` that counts occurrences of specific values in a variable

The only additional Analyze Function we need here is `count_values` that counts occurrences of 
specific values in a variable. It depends on a generic function `s_count_value` that works with
different default `value`s. The other arguments are inherited from `s_summary.logical` which will 
be used as the workhorse inside (which is in turn very similar to `s_summary.factor`, so we might be 
able to create some helper function that serves both of these `s_summary` methods).

```{r s_summary.logical}
getS3method("s_summary", "logical")
```

```{r s_count_value}
s_count_values
s_count_values(factor(c("N", "Y", NA, "Y")))
s_count_values(factor(c("N", "Y", NA, "Y")), na.rm = FALSE)
```

### Analyze Function

```{r count_values}
count_values
```

### Test 

```{r simple variant test}
basic_table() %>%
  split_cols_by("ARM", split_fun = add_overall_level("All Patients", first = FALSE)) %>%
  add_colcounts() %>%
  count_values("COMPSTUD", values = "Y", .labels = c(count_fraction = "Completed Study")) %>%
  split_rows_by("DISSTDFL", split_fun = keep_split_levels("Y")) %>%
  summarize_row_groups(label_fstr = "Discontinued Study") %>%
  summarize_vars("STDDRS", .stats = "count_fraction") %>%
  build_table(adsl0)
```

## Variant 2: With grouping of reasons

We want this:

```markdown
                                A: Drug X             B: Placebo          C: Combination         All Patients 
                                  (N=134)               (N=134)               (N=132)               (N=400)    
---------------------------------------------------------------------------------------------------------------
Received treatment               134 (100%)            134 (100%)            132 (100%)            400 (100%)  
Discontinued treatment          71 (52.99%)           62 (46.27%)           75 (56.82%)            208 (52%)   
  Safety                        27 (20.15%)           27 (20.15%)            26 (19.7%)             80 (20%)   
    ADVERSE EVENT                12 (8.96%)           14 (10.45%)           15 (11.36%)           41 (10.25%)  
    PHYSICIAN DECISION          15 (11.19%)            13 (9.7%)             11 (8.33%)            39 (9.75%)  
  Other                         44 (32.84%)           35 (26.12%)           49 (37.12%)            128 (32%)   
    LACK OF EFFICACY             13 (9.7%)            19 (14.18%)           21 (15.91%)           53 (13.25%)  
    OTHER                       15 (11.19%)            10 (7.46%)            12 (9.09%)            37 (9.25%)  
    PROGRESSIVE DISEASE         16 (11.94%)            6 (4.48%)            16 (12.12%)            38 (9.5%) 
```

Let's try:

```{r subgrouping}
basic_table() %>%
  split_cols_by("ARM", split_fun = add_overall_level("All Patients", first = FALSE)) %>%
  add_colcounts() %>%
  count_values("GOTTRT", values = "Y", .labels = c(count_fraction = "Received treatment")) %>%
  split_rows_by("DISTRTFL", split_fun = keep_split_levels("Y")) %>%
  summarize_row_groups(label_fstr = "Discontinued treatment") %>%
  split_rows_by("DRSCAT", split_fun = reorder_split_levels(c("Safety", "Other"))) %>%
  summarize_row_groups() %>%
  summarize_vars("TRTDRS", .stats = "count_fraction") %>%
  build_table(adsl0) %>%
  prune_table()
```

## Variant 3: Adding other optional rows

Of course we can also add more `count_values` rows on top. 

```{r optional rows}
basic_table() %>%
  split_cols_by("ARM", split_fun = add_overall_level("All Patients", first = FALSE)) %>%
  add_colcounts() %>%
  count_values("COMPSTUD", values = "Y", .labels = c(count_fraction = "Completed Study")) %>%
  count_values("STUDONS", values = "Alive: In Follow-up", .labels = c(count_fraction = "Alive: In Follow-up")) %>%
  split_rows_by("DISSTDFL", split_fun = keep_split_levels("Y"), nested = FALSE) %>%
  summarize_row_groups(label_fstr = "Discontinued Study") %>%
  summarize_vars("STDDRS", .stats = "count_fraction") %>%
  build_table(adsl0)
```
