---
title: "Introduction to tern"
output:
    rmarkdown::html_document:
        theme: "spacelab"
        highlight: "kate"
        toc: true
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Introduction to tern}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
editor_options:
    markdown: 
        wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>"
)
```

## Introduction to tern

---------

This vignette shows the `tern` package, its general purpose and syntax.  
The `tern` R package contains analytical functions for creating tables and graphs useful for clinical trials and other statistical analysis. 
The main focus is on the clinical trial reporting tables. 
Nevertheless, the graphs related to the clinical trials are a valuable supplement.
The core functionality for tabulation is built on top of the more general purpose `rtables` package. 

**It is strongly recommended that you start by reading the "Introduction to rtables" vignette to get familiar with the concept of `rtables`.**

---------

## Common Clinical Trials Analyses

The package provides a large range of functionality to create tables and graphs used for clinical trial and other statistical analysis.

`rtables` tabulation extended by clinical trials specific functions:

- demographics
- unique patients
- exposure across patients
- change from baseline for parameters
- statistical model fits: MMRM, logistic regression, Cox regression, ...
- ...

`rtables` tabulation helper functions:

- pre-processing
- conversions and transformations
- ...

data visualizations connected with clinical trials:

- Kaplan-Meier plots
- forest plots
- line plots
- ...

data visualizations helper functions:

- arrange/stack multiple graphs
- decorate graphs like add titles, footnotes, page number, ...
- ...


The reference of `tern` functions is available on [the tern website functions reference](https://insightsengineering.github.io/tern/reference/index.html).

---------

## Analyze Functions for `rtables`

Analyze functions are used in combination with the `rtables` layout functions, in the pipeline which creates the `rtables` table.
They apply some statistical logic to the layout of the `rtables` table.
The table layout is materialized with the `rtables::build_table` function and the data.

The `tern` analyze functions are wrappers around `rtables::analyze` function, they offer various methods useful from the perspective of clinical trials and other statistical projects.

Examples of the `tern` analyze functions are `tern::count_occurrences`, `tern::summarize_ancova` or `tern::summarize_vars`. 
As there is no one prefix to identify all `tern` analyze functions it is recommended to use the reference subsection on [the tern website](https://insightsengineering.github.io/tern/reference/index.html).

In the `rtables` code below we first described the table and assigned that description to a variable `l`. We then built the table using the actual data with `rtables::build_table`. The description of a table is called a table **layout**. The **analyze instruction** adds to the layout that the `ARM` variable should be analyzed with the `mean` analysis function and the result should be rounded to 1 decimal place.
Hence, a **layout** is “pre-data”, that is, it’s a description of **how to build a table once we get data**.

```{r, message=FALSE}
library(tern)
library(dplyr)

adrs <- ex_adrs
```

Defining the table layout with a pure `rtables` code.

```{r}
# Create table layout pure rtables
l <- basic_table() %>%
  split_cols_by(var = "ARM") %>%
  split_rows_by(var = "AVISIT") %>%
  analyze(vars = "AVAL", mean, format = "xx.x")
```

Below the only `tern` function is the `summarize_vars` which replace the `rtables::analyze` one.

```{r}
# Create table layout with tern summarize_vars analyze function
l2 <- basic_table() %>%
  split_cols_by(var = "ARM") %>%
  split_rows_by(var = "AVISIT") %>%
  summarize_vars(vars = "AVAL", .formats = c("mean_sd" = "(xx.xx, xx.xx)"))
```

```{r}
# Apply table layout to data and produce `rtables` object
build_table(l, df = adrs)
build_table(l2, df = adrs)
```

We could find out that `tern` offers more advanced analysis with only a one code line updated compared to the pure `rtables` code. 

**More examples with tabulation analyze functions are presented in the `tern tabulation` vignette.**

## Clinical Trials Visualizations

Clinical trails related plots complement the rich palette of `tern` tabulation analysis functions.
Thus the `tern` package delivers a full-featured tool for clinical trial reporting.
The `tern` plot functions return `ggplot2` or `gTree` objects, the latter is returned when a table is attached to the plot.

```{r}
adsl <- ex_adsl
adlb <- ex_adlb
adlb <- dplyr::filter(adlb, PARAMCD == "ALT", AVISIT != "SCREENING")
```

Line plot without a table applied by the `tern::g_lineplot` function.

```{r}
# Mean with CI
tern::g_lineplot(adlb, adsl, subtitle = "Laboratory Test:")
```

Line plot with a table applied by the `tern::g_lineplot` function.

```{r, fig.height = 10}
# Mean with CI, table and customized confidence level
tern::g_lineplot(
  adlb,
  adsl,
  table = c("n", "mean", "mean_ci"),
  title = "Plot of Mean and 80% Confidence Limits by Visit"
)
```

The first plot is of the `ggplot2` and the second `gTree` type, as the latter contains the table.

The `tern` functions used for plot generation are mostly `g_` prefixed.
All `tern` plot functions are listed on [the tern website functions reference](https://insightsengineering.github.io/tern/reference/index.html).

## Interactive Apps

Most of `tern` outputs could be easily accommodated into `shiny` apps. 
We are recommending to apply `tern` outputs into the `teal` apps. 
The [`teal` package](https://insightsengineering.github.io/teal/) is a shiny-based interactive exploration framework for analyzing data.
`teal` modules with `tern` outputs are available in the [`teal.modules.clinical` package](https://insightsengineering.github.io/teal.modules.clinical/).

## Summary 

In summary, `tern` contains many additional functions for creating tables, listing and graphs used in clinical trials and other statistical analysis. The design of the package gives users a lot of flexibility to meet the analysis needs in a regulatory or exploratory reporting context.

**For more information please visit [the tern website](https://insightsengineering.github.io/tern).**
