---
title: "rtables: Reporting Tables"
author: "Adrian Waddell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

`rtable` is useful to create tables where multiple values are stored in a cell 
with a particular format. An `rtable` object has access to the individual raw 
values and prints the data with formatting rules such as rounding and pasting.
`rtables` also allow for merging cells within a row.

The main functions currently associated with `rtable`s are `rtable`, `rrow`,
`rcell`, and `as_html`.

Consider the foolowing example:

```{r}
library(teal.oncology)

tbl <- rtable(
  col.names = c("Treatement\nN=100", "Comparison\nN=300"),
  format = "xx (xx.xx%)",
  rrow("Response", c(104, .2), c(100, .4)),
  rrow("Non-Response", c(23, .4), c(43, .5)),
  rrow(),
  rrow("this is a very long section header"),
  rrow("HR", rcell(3.23, "xx.xx", colspan = 2)),
  rrow("95% CI", indent = 2, rcell(c(2.8, 4.4), format = "(xx.x, xx.x)", colspan = 2))
)

tbl  
```


Before we go into explaining the individual components used to create this table
we continue with the html conversio of the `rtbale` object:


```{r}
as_html(tbl, width = "80%")
```

Next, the `[` operator lets you access the cell content.

```{r}
tbl[1, 1]
```

and to format that cell run `format_rcell(tbl[1,1])`=`r format_rcell(tbl[1,1])`.

Note that `tbl[6, 1]` and `tbl[6, 2]` display both the same `rcell` because of
the colspan.

# Viewer

The `Viewer` functions takes an `rtable` object and displays it with bootstrap
styling in RStudio's Viewer pane.

```{r, eval = FALSE}
Viewer(tbl)
```


# Formats


# Converting table objects

You can also convert table objects to rtables:

```{r}
as.rtable(table(iris$Species))
```

or 

```{r}
print(as.rtable(with(mtcars, table(cyl, hp))), gap = 1)
```