---
title: "Introduction to tern"
date: "2022-03-09"
output:
    rmarkdown::html_document:
        theme: "spacelab"
        highlight: "kate"
        toc: true
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Introduction to tern}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
editor_options:
    markdown: 
        wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction to tern

---------

This vignette shows the general purpose and syntax of the `tern` R package.  
The `tern` R package contains analytical functions for creating tables and graphs useful for clinical trials and other statistical analysis. 
The main focus is on the clinical trial reporting tables but the graphs related to the clinical trials are also valuable.
The core functionality for tabulation is built on top of the more general purpose `rtables` package. 

[**It is strongly recommended that you start by reading the "Introduction to `rtables`" vignette to get familiar with the concept of `rtables`.**](https://roche.github.io/rtables/articles/introduction.html)

---------

## Common Clinical Trials Analyses

The package provides a large range of functionality to create tables and graphs used for clinical trial and other statistical analysis.

`rtables` tabulation extended by clinical trials specific functions:

- demographics
- unique patients
- exposure across patients
- change from baseline for parameters
- statistical model fits: MMRM, logistic regression, Cox regression, ...
- ...

`rtables` tabulation helper functions:

- pre-processing
- conversions and transformations
- ...

data visualizations connected with clinical trials:

- Kaplan-Meier plots
- forest plots
- line plots
- ...

data visualizations helper functions:

- arrange/stack multiple graphs
- embellishing graphs/tables with metadata and details, such as adding titles, footnotes, page number, etc.
- ...


The reference of `tern` functions is available on [the tern website functions reference](https://insightsengineering.github.io/tern/reference/index.html).

---------

## Analytical Functions for `rtables`

Analytical functions are used in combination with other `rtables` layout functions, in the pipeline which creates the `rtables` table.
They apply some statistical logic to the layout of the `rtables` table.
The table layout is materialized with the `rtables::build_table` function and the data.

The `tern` analytical functions are wrappers around the `rtables::analyze` function; they offer various methods useful from the perspective of clinical trials and other statistical projects.

Examples of the `tern` analytical functions are `tern::count_occurrences`, `tern::summarize_ancova` and `tern::summarize_vars`. 
As there is no one prefix to identify all `tern` analytical functions it is recommended to use the reference subsection on [the tern website](https://insightsengineering.github.io/tern/reference/index.html).

In the `rtables` code below we first describe the two tables and assign the descriptions to the variables `lyt` and `lyt2`. We then built the tables using the actual data with `rtables::build_table`. The description of a table is called a table **layout**. The **analyze instruction** adds to the layout that the `ARM` variable should be analyzed with the `mean` analysis function and the result should be rounded to 1 decimal place.
Hence, a **layout** is “pre-data”; that is, it’s a description of **how to build a table once we get data**.

```{r, message=FALSE}
library(tern)
library(dplyr)
```

Defining the table layout with a pure `rtables` code.

```{r}
# Create table layout pure rtables
lyt <- rtables::basic_table() %>%
  rtables::split_cols_by(var = "ARM") %>%
  rtables::split_rows_by(var = "AVISIT") %>%
  rtables::analyze(vars = "AVAL", mean, format = "xx.x")
```

Below the only `tern` function is `summarize_vars` which replaces the `rtables::analyze` function above.

```{r}
# Create table layout with tern summarize_vars analyze function
lyt2 <- rtables::basic_table() %>%
  rtables::split_cols_by(var = "ARM") %>%
  rtables::split_rows_by(var = "AVISIT") %>%
  tern::summarize_vars(vars = "AVAL", .formats = c("mean_sd" = "(xx.xx, xx.xx)"))
```

```{r}
# Apply table layout to data and produce `rtables` object

adrs <- formatters::ex_adrs

rtables::build_table(lyt, df = adrs)
rtables::build_table(lyt2, df = adrs)
```

We see that `tern` offers advanced analysis by extending `rtables` function calls with only one additional function call. 

**More examples with tabulation analyze functions are presented in the `Tabulation` vignette.**

## Clinical Trials Visualizations

Clinical trial related plots complement the rich palette of `tern` tabulation analysis functions.
Thus the `tern` package delivers a full-featured tool for clinical trial reporting.
The `tern` plot functions return `ggplot2` or `gTree` objects, the latter is returned when a table is attached to the plot.

```{r}
adsl <- formatters::ex_adsl
adlb <- formatters::ex_adlb
adlb <- dplyr::filter(adlb, PARAMCD == "ALT", AVISIT != "SCREENING")
```

Line plot without a table generated by the `tern::g_lineplot` function.

```{r}
# Mean with CI
tern::g_lineplot(adlb, adsl, subtitle = "Laboratory Test:")
```

Line plot with a table generated by the `tern::g_lineplot` function.

```{r, fig.height=10, fig.width=8}
# Mean with CI, table and customized confidence level
tern::g_lineplot(
  adlb,
  adsl,
  table = c("n", "mean", "mean_ci"),
  title = "Plot of Mean and 80% Confidence Limits by Visit"
)
```

The first plot is a `ggplot2` object and the second plot is a `gTree` object, as the latter contains the table.
The second plot has to be properly resized to get a clear and readable table content.

The `tern` functions used for plot generation are mostly `g_` prefixed.
All `tern` plot functions are listed on [the tern website functions reference](https://insightsengineering.github.io/tern/reference/index.html).

## Interactive Apps

Most of `tern` outputs could be easily accommodated into `shiny` apps. 
We recommend applying `tern` outputs into `teal` apps. 
The [`teal` package](https://insightsengineering.github.io/teal/) is a shiny-based interactive exploration framework for analyzing data.
`teal` shiny apps with `tern` outputs are available in the [`teal.modules.clinical` package](https://insightsengineering.github.io/teal.modules.clinical/).

## Summary 

In summary, `tern` contains many additional functions for creating tables, listing and graphs used in clinical trials and other statistical analyses. The design of the package gives users a lot of flexibility to meet the analysis needs in a regulatory or exploratory reporting context.

**For more information please explore [the tern website](https://insightsengineering.github.io/tern).**
