---
title: "Introduction to tern"
date: "2022-03-09"
output:
    rmarkdown::html_document:
        theme: "spacelab"
        highlight: "kate"
        toc: true
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Introduction to tern}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
editor_options:
    markdown:
        wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows the general purpose and syntax of the `tern` R package.
The `tern` R package contains analytical functions for creating tables and graphs useful for clinical trials and other statistical analysis.
The main focus is on the clinical trial reporting tables but the graphs related to the clinical trials are also valuable.
The core functionality for tabulation is built on top of the more general purpose `rtables` package.

[**It is strongly recommended that you start by reading the "Introduction to `rtables`" vignette to get familiar with the concept of `rtables`.**](https://pharmaverse.github.io/rtables/latest-tag/articles/rtables.html)

---------

## Common Clinical Trials Analyses

The package provides a large range of functionality to create tables and graphs used for clinical trial and other statistical analysis.

`rtables` tabulation extended by clinical trials specific functions:

- demographics
- unique patients
- exposure across patients
- change from baseline for parameters
- statistical model fits: MMRM, logistic regression, Cox regression, ...
- ...

`rtables` tabulation helper functions:

- pre-processing
- conversions and transformations
- ...

data visualizations connected with clinical trials:

- Kaplan-Meier plots
- forest plots
- line plots
- ...

data visualizations helper functions:

- arrange/stack multiple graphs
- embellishing graphs/tables with metadata and details, such as adding titles, footnotes, page number, etc.
- ...


The reference of `tern` functions is available on [the tern website functions reference](https://pharmaverse.github.io/tern/latest-tag/reference/index.html).

---------

## Analytical Functions for `rtables`

Analytical functions are used in combination with other `rtables` layout functions, in the pipeline which creates the `rtables` table.
They apply some statistical logic to the layout of the `rtables` table.
The table layout is materialized with the `rtables::build_table` function and the data.

The `tern` analytical functions are wrappers around the `rtables::analyze` function; they offer various methods useful from the perspective of clinical trials and other statistical projects.

Examples of the `tern` analytical functions are `count_occurrences`, `summarize_ancova` and `analyze_vars`.
As there is no one prefix to identify all `tern` analytical functions it is recommended to use the reference subsection on [the tern website](https://pharmaverse.github.io/tern/latest-tag/reference/index.html).

In the `rtables` code below we first describe the two tables and assign the descriptions to the variables `lyt` and `lyt2`. We then built the tables using the actual data with `rtables::build_table`. The description of a table is called a table **layout**. The **analyze instruction** adds to the layout that the `ARM` variable should be analyzed with the `mean` analysis function and the result should be rounded to 1 decimal place.
Hence, a **layout** is “pre-data”; that is, it’s a description of **how to build a table once we get data**.

```{r, message=FALSE}
library(tern)
library(dplyr)
```

Defining the table layout with a pure `rtables` code:

```{r}
# Create table layout pure rtables
lyt <- rtables::basic_table() |>
  rtables::split_cols_by(var = "ARM") |>
  rtables::split_rows_by(var = "AVISIT") |>
  rtables::analyze(vars = "AVAL", mean, format = "xx.x")
```

Below, the only `tern` function used is `analyze_vars` which replaces the `rtables::analyze` function used above.

```{r}
# Create table layout with tern analyze_vars analyze function
lyt2 <- rtables::basic_table() |>
  rtables::split_cols_by(var = "ARM") |>
  rtables::split_rows_by(var = "AVISIT") |>
  analyze_vars(vars = "AVAL", .formats = c("mean_sd" = "(xx.xx, xx.xx)"))
```

```{r}
# Apply table layout to data and produce `rtables` object

adrs <- formatters::ex_adrs

rtables::build_table(lyt, df = adrs)
rtables::build_table(lyt2, df = adrs)
```

We see that `tern` offers advanced analysis by extending `rtables` function calls with only one additional function call.

**More examples with tabulation analyze functions are presented in the `Tabulation` vignette.**

## Clinical Trial Visualizations

Clinical trial related plots complement the rich palette of `tern` tabulation analysis functions.
Thus the `tern` package delivers a full-featured tool for clinical trial reporting.
The `tern` plot functions return graphs as `ggplot2` objects.

```{r}
adsl <- formatters::ex_adsl
adlb <- formatters::ex_adlb
adlb <- dplyr::filter(adlb, PARAMCD == "ALT", AVISIT != "SCREENING")
```

The `nestcolor` package can be loaded in to apply the standardized NEST color palette to all `tern` plots.

```{r}
library(nestcolor)
```

Line plot without a table generated by the `g_lineplot` function.

```{r, fig.alt='Basic line plot'}
# Mean with CI
g_lineplot(adlb, adsl, subtitle = "Laboratory Test:")
```

Line plot with a table generated by the `g_lineplot` function.

```{r, fig.height=10, fig.width=8, fig.alt='Line plot with table'}
# Mean with CI, table, and customized confidence level
g_lineplot(
  adlb,
  adsl,
  table = c("n", "mean", "mean_ci"),
  title = "Plot of Mean and 80% Confidence Limits by Visit"
)
```

### Line plot customization

The `g_lineplot()` function provides several arguments for customizing
the summary statistics and appearance of a clinical trial line plot.

The `mid` argument specifies the summary statistic displayed at each
time point. For example, the default mean can be changed to the median.
The corresponding confidence interval and whisker variables should
also be specified.

```{r, fig.alt='Median line plot'}
g_lineplot(
  adlb,
  adsl,
  mid = "median",
  interval = "median_ci",
  whiskers = c("median_ci_lwr", "median_ci_upr"),
  title = "Plot of Median and 95% Confidence Limits by Visit"
)
```
The confidence level can be customized using the control argument.
For example, the following plot uses an 80% confidence interval.

```{r, fig.height=10, fig.width=8, fig.alt='Line plot with customized confidence level'}
g_lineplot(
  adlb,
  adsl,
  table = c("n", "mean", "mean_ci"),
  control = control_analyze_vars(conf_level = 0.80),
  title = "Plot of Mean and 80% Confidence Limits by Visit"
)
```
Other commonly used arguments include mid_type for controlling how
the midpoint is displayed, position for controlling the position of
groups, and legend_position for changing the legend location. The
col and linetype arguments can also be used to customize the
appearance of the plotted groups.

### Kaplan-Meier plot

A Kaplan-Meier plot can be generated using `g_km()`. Kaplan-Meier
plots are commonly used in clinical trials to visualize time-to-event
outcomes, such as overall survival or progression-free survival.

The function requires a data set containing the time-to-event variable,
an event indicator, and a treatment group variable. These variables are
specified through the `variables` argument.

```{r, fig.alt='Kaplan-Meier plot'}
adtte <- formatters::ex_adtte |>
  dplyr::filter(PARAMCD == "OS") |>
  dplyr::mutate(is_event = CNSR == 0)

variables <- list(
  tte = "AVAL",
  is_event = "is_event",
  arm = "ARMCD"
)

g_km(
  df = adtte,
  variables = variables
)
```

In the example above, PARAMCD == "OS" selects the overall survival parameter from the example time-to-event data. The is_event variable is derived from the censoring variable, with CNSR == 0 indicating that the event has occurred.
The elements of variables specify the variables required by g_km():
- tte specifies the time-to-event duration variable.
- is_event specifies whether an event occurred (TRUE) or the
  observation was censored (FALSE).
- arm specifies the treatment group variable.
The resulting plot displays the Kaplan-Meier survival curves for each treatment group. By default, censored observations are shown on the curves, and the plot can include a median survival summary and a patients-at-risk table.
The Kaplan-Meier plot can be customized using arguments such as yval, ci_ribbon, max_time, annot_at_risk, and annot_surv_med.

```{r, fig.alt='Customized Kaplan-Meier plot'}
g_km(
  df = adtte,
  variables = variables,
  yval = "Failure",
  ci_ribbon = TRUE,
  max_time = 500,
  annot_at_risk = TRUE,
  annot_surv_med = TRUE
)
```

In this example:
- yval = "Failure" displays failure probability instead of survival probability.
- ci_ribbon = TRUE adds confidence interval ribbons around the Kaplan-Meier curves.
- max_time = 500 limits the x-axis to 500 days.
- annot_at_risk = TRUE displays the number of patients at risk at the time points shown on the x-axis.
- annot_surv_med = TRUE adds the estimated median survival time for each treatment group.
Other arguments can be used to customize the appearance and statistical details of the Kaplan-Meier plot. For example, `control_surv` can be used
to specify confidence interval settings, while `col`, `lty`, and `lwd` can be used to customize the colors, line types, and line widths of the
treatment-group curves. The `xticks` argument can be used to control the positions of the x-axis tick marks.

The `max_time` argument should be selected according to the follow-up period of the data. When `max_time` is set substantially beyond the
observed follow-up period, the patients-at-risk table may contain zero patients at later time points.

### Forest plot

A forest plot can be generated using `g_forest()`. Forest plots are
commonly used in clinical trials to display treatment-effect estimates
and their confidence intervals across patient subgroups.

The `g_forest()` function uses a table containing an effect estimate and
its corresponding confidence interval. For ratio measures such as odds
ratios, the plot is displayed on a logarithmic x-axis with a reference
line at 1.

The following example prepares response subgroup data and creates a
table containing the total number of patients, odds ratio, and
confidence interval for each subgroup.

```{r, fig.alt='Forest plot'}
adrs <- tern_ex_adrs
n_records <- 20

adrs_labels <- formatters::var_labels(adrs, fill = TRUE)

adrs <- adrs |>
  dplyr::filter(PARAMCD == "BESRSPI") |>
  dplyr::filter(ARM %in% c("A: Drug X", "B: Placebo")) |>
  dplyr::slice(seq_len(n_records)) |>
  droplevels() |>
  dplyr::mutate(
    ARM = forcats::fct_relevel(ARM, "B: Placebo"),
    rsp = AVALC == "CR"
  )

formatters::var_labels(adrs) <- c(adrs_labels, "Response")

df <- extract_rsp_subgroups(
  variables = list(
    rsp = "rsp",
    arm = "ARM",
    subgroups = c("SEX", "STRATA2")
  ),
  data = adrs
)

tbl_or <- basic_table() |>
  tabulate_rsp_subgroups(
    df,
    vars = c("n_tot", "or", "ci")
  )

g_forest(tbl_or)
```

In this example, extract_rsp_subgroups() prepares the response subgroup data. The tabulate_rsp_subgroups() function then creates a table containing the selected statistics. The n_tot, or, and ci statistics are used because they provide the total number of patients, the odds ratio, and its confidence interval required for the forest plot.
The forest plot can be customized using arguments such as forest_header, vline, xlim, x_at, and logx.

```{r, fig.alt='Customized forest plot'}
g_forest(
  tbl_or,
  forest_header = c("Placebo\nBetter", "Drug X\nBetter"),
  vline = 1,
  xlim = c(0.1, 10),
  x_at = c(0.1, 0.5, 1, 2, 10),
  logx = TRUE
)
```

In this example:
- forest_header specifies the labels displayed on either side of the
  reference line.
- vline = 1 adds a reference line at the null value for the odds ratio.
- xlim specifies the limits of the x-axis.
- x_at specifies the positions of the x-axis tick marks.
- logx = TRUE displays the x-axis on a logarithmic scale.
For ratio measures such as odds ratios, relative risks, and hazard ratios, the null value is generally 1. Therefore, a reference line at 1 is commonly used when displaying these treatment-effect estimates.
Additional arguments such as col, font_size, and rel_width_forest can be used to customize the appearance and layout of the forest plot.

All `tern` functions used for plot generation are `g_` prefixed and are listed on [the tern website functions reference](https://pharmaverse.github.io/tern/latest-tag/reference/index.html#graphs).

## Interactive Apps

Most `tern` outputs can be easily converted into `shiny` apps. We recommend building apps using the [`teal` package](https://insightsengineering.github.io/teal/latest-tag/), a shiny-based interactive exploration framework for analyzing data.
A variety of pre-made `teal` shiny apps for `tern` outputs are available in the [`teal.modules.clinical` package](https://insightsengineering.github.io/teal.modules.clinical/latest-tag/).

## Summary

In summary, `tern` contains many additional functions for creating tables, listings, and graphs used in clinical trials and other statistical analyses. The design of the package gives users the flexibility to meet the analysis needs in both regulatory and exploratory reporting contexts.

**For more information please explore [the tern website](https://pharmaverse.github.io/tern/latest-tag/).**
