---
title: "Unconditional Exact Confidence Interval for Difference in Proportions"
date: "2026-05-15"
output:
    rmarkdown::html_document:
        theme: "spacelab"
        highlight: "kate"
        toc: true
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Unconditional Exact Confidence Interval for Difference in Proportions}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
bibliography: ../inst/REFERENCES.bib
editor_options:
    markdown:
        wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette explains the construction of the unconditional exact confidence interval for the difference in proportions using the `estimate_proportion_diff()` function with `method = "uncond_exact_diff"` from the `tern` package. This method is particularly useful when dealing with small sample sizes or when the assumptions of other methods (like the normal approximation) are not met.

## Definition

The definition of the exact unconditional confidence limits is as follows.
Without loss of generality, we define $p_1$ as the risk (proportion) to be in column 1 for row 1, and $p_2$ as the risk (proportion) to be in column 1 for row 2. The risk difference is defined as the difference between the row 1 and row 2 risks
(proportions), $d = p_1 - p_2$, and $n_1$ and $n_2$ denote the row totals of the
$2 \times 2$ table. The probability function for a given $2 \times 2$ table

$$
\begin{array}{c|cc|c}
& \text{Column 1} & \text{Column 2} & \text{Row Total} \\
\hline
\text{Row 1} & n_{11} & n_{12} & n_1 \\
\text{Row 2} & n_{21} & n_{22} & n_2 \\
\hline
\end{array}
$$

can then be expressed in terms of the table cell frequencies, the risk difference, and the nuisance parameter $p_2$ as:

$$
f(n_{11}, n_{21}; n_1, n_2, d, p_2)
= \binom{n_1}{n_{11}} (d + p_2)^{n_{11}} (1 - d - p_2)^{n_1 - n_{11}}
\cdot \binom{n_2}{n_{21}} p_2^{n_{21}} (1 - p_2)^{n_2 - n_{21}}.
$$

The exact unconditional approach fixes the row margins $n_1$ and $n_2$ of the $2 \times 2$ table and eliminates the nuisance parameter $p_2$ by using the maximum p-value (worst-case scenario) over all possible values of $p_2$ [@SantnerSnell1980].
It then computes the confidence limits by inverting two separate one-sided exact tests that are based on the unstandardized risk difference $d$.

For a given $100(1 - \alpha/2)\%$ confidence level, the confidence limits for the risk difference $d$ are computed as

$$
d_L = \sup \{ d_* : P_U(d_*) > \alpha/2 \},
\qquad
d_U = \inf \{ d_* : P_L(d_*) > \alpha/2 \}.
$$

where

$$
P_U(d_*) = \sup_{p_2}
\left(
    \sum_{A,\,T(a) \ge t_0}
    f(n_{11}, n_{21}; n_1, n_2, d_*, p_2)
\right)
$$

and

$$
P_L(d_*) = \sup_{p_2}
\left(
    \sum_{A,\,T(a) \le t_0}
    f(n_{11}, n_{21}; n_1, n_2, d_*, p_2)
\right).
$$

The set $A$ includes all $2 \times 2$ tables in which the row sums are $n_1$ and
$n_2$, as these are assumed fixed as explained above. $T(a)$ denotes the value of the test statistic, here the unstandardized risk
difference, for table $a$ in $A$, which depends on the counts $n_{11}(a)$ and $n_{21}(a)$ in the first column of the table, and is defined as:

$$
T(a) = \frac{n_{11}(a)}{n_1} - \frac{n_{21}(a)}{n_2}.
$$

$t_0$ is the value of the test statistic for the observed table.

To compute $P_U(d_*)$, the sum includes probabilities of those tables
for which $T(a) \ge t_0$. For a fixed value of $d_*$, $P_U(d_*)$ is defined as
the maximum sum over all possible values of $p_2$.

Similarly, to compute $P_L(d_*)$, the sum includes probabilities of those tables
for which $T(a) \le t_0$. For a fixed value of $d_*$, $P_L(d_*)$ is defined as
the maximum sum over all possible values of $p_2$.

Finally, the confidence limits are defined as the supremum of $d_*$ values for which $P_U(d_*)$ is greater than $\alpha/2$ for the lower limit, and the infimum of $d_*$ values for which $P_L(d_*)$ is greater than $\alpha/2$ for the upper limit.

## Prototype

The following is a readable prototype implementation that follows the
definition step by step: table enumeration ($A$), test statistic
$T(a)$, table probability $f(\cdot)$, worst-case one-sided p-values
$P_U(d_*)$ and $P_L(d_*)$, and inversion to confidence limits.

```{r, eval = FALSE}
# Step 1: Enumerate all tables in A with fixed row margins n1 and n2.
enumerate_tables_fixed_margins <- function(n1, n2) {
  expand.grid(n11 = 0:n1, n21 = 0:n2)
}

# Step 2: Compute T(a) = n11 / n1 - n21 / n2 for each table a in A.
risk_difference_statistic <- function(n11, n21, n1, n2) {
  n11 / n1 - n21 / n2
}

# Step 3: Compute f(n11, n21; n1, n2, d, p2).
table_probability_f <- function(n11, n21, n1, n2, d, p2) {
  p1 <- d + p2
  dbinom(n11, size = n1, prob = p1) * dbinom(n21, size = n2, prob = p2)
}

# Step 4: For fixed d, determine valid nuisance-parameter range for p2
# so that p2 in [0, 1] and p1 = d + p2 in [0, 1].
valid_p2_interval <- function(d) {
  lower <- max(0, -d)
  upper <- min(1, 1 - d)
  c(lower, upper)
}

# Step 5: Compute worst-case one-sided tail probability:
#   P_U(d*) = sup_p2 sum_{T(a) >= t0} f(...)
#   P_L(d*) = sup_p2 sum_{T(a) <= t0} f(...)
worst_case_tail_probability <- function(d_star, n11_obs, n21_obs, n1, n2,
                                        tail = c("upper", "lower")) {
  tail <- match.arg(tail)
  tables <- enumerate_tables_fixed_margins(n1, n2)
  t_values <- risk_difference_statistic(
    n11 = tables$n11, n21 = tables$n21, n1 = n1, n2 = n2
  )
  t0 <- risk_difference_statistic(n11_obs, n21_obs, n1, n2)

  include_table <- if (tail == "upper") {
    t_values >= t0
  } else {
    t_values <= t0
  }

  objective <- function(p2) {
    probs <- table_probability_f(
      n11 = tables$n11,
      n21 = tables$n21,
      n1 = n1,
      n2 = n2,
      d = d_star,
      p2 = p2
    )
    sum(probs[include_table])
  }

  interval <- valid_p2_interval(d_star)
  if (interval[1] == interval[2]) {
    return(objective(interval[1]))
  }

  optimize(
    f = objective,
    interval = interval,
    maximum = TRUE
  )$objective
}

# Step 6: Invert one-sided tests to obtain the two-sided
# 100 * (1 - alpha)% CI for d = p1 - p2.
uncond_exact_diff_ci_prototype <- function(n11_obs, n21_obs, n1, n2,
                                           alpha = 0.05,
                                           d_grid = seq(-1, 1, by = 0.001)) {
  cutoff <- alpha / 2

  p_upper <- vapply(
    d_grid,
    FUN.VALUE = numeric(1),
    FUN = function(d_star) {
      worst_case_tail_probability(
        d_star = d_star,
        n11_obs = n11_obs,
        n21_obs = n21_obs,
        n1 = n1,
        n2 = n2,
        tail = "upper"
      )
    }
  )

  p_lower <- vapply(
    d_grid,
    FUN.VALUE = numeric(1),
    FUN = function(d_star) {
      worst_case_tail_probability(
        d_star = d_star,
        n11_obs = n11_obs,
        n21_obs = n21_obs,
        n1 = n1,
        n2 = n2,
        tail = "lower"
      )
    }
  )

  lower_candidates <- d_grid[p_upper > cutoff]
  upper_candidates <- d_grid[p_lower > cutoff]

  d_L <- if (length(lower_candidates)) max(lower_candidates) else NA_real_
  d_U <- if (length(upper_candidates)) min(upper_candidates) else NA_real_

  list(
    estimate = n11_obs / n1 - n21_obs / n2,
    conf_level = 1 - alpha,
    conf_int = c(lower = d_L, upper = d_U),
    details = list(
      d_grid = d_grid,
      p_upper = p_upper,
      p_lower = p_lower,
      cutoff = cutoff
    )
  )
}

# Example: observed first-column counts n11 = 7 of n1 = 20,
# and n21 = 3 of n2 = 18.
prototype_result <- uncond_exact_diff_ci_prototype(
  n11_obs = 7,
  n21_obs = 3,
  n1 = 20,
  n2 = 18,
  alpha = 0.05
)

prototype_result$estimate
prototype_result$conf_int
```

## References
