---
title: "Design: Kaplan-Meier Plot (`KMG01`)"
author: "Francois Collin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    theme: journal
    highlight: zenburn
vignette: >
  %\VignetteIndexEntry{Design of KMG01}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../inst/REFERENCES.bib

---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "svg"
)
```

Introduction
============

+ Existing TLG [entry](https://nest-docs.kubemea.science.roche.com/releases/2020_10_20/embedded/agile-R/tlg_catalog/graphs/kmg01/)
+ Existing
[prototype](https://pages.github.roche.com/collinf1/RGraphics/RGraphics_grid/RGraphics_grid.html) 
following
[issue 499](https://github.roche.com/NEST/tern/issues/499).


Specification: the function `g_km` is the current version of the API and the
new prototype should accept compatible arguments.


Method
======

The principle consists in using the acknowledge high level functions from
`ggplot2` to render the Kaplan-Meier curve. Then, after decomposition of
the graphics, the elements are arrange in a new layout with predefined positions
for the `ggplot` elements (y or x axis, graphic and legends) and add annotation
tables:

- the survival median table overlay optionally on the top right corner above 
the curve.
- the patient at risk table with a constraint as the columns must be aligned
to the curve y-axis.


Result
======


ggplot2 decomposition {.tabset}
---------------------

`ggplot2` offers a comfortable interface to draw the the curve.

The idea:

1. the survival model is fitted and the data extracted: `h_data_plot`.
2. the x ticks are estimated to be used in the graph and the patient-at-risk
table.
3. the graphics is obtained with `ggplot` ...
4. ... and then decomposed into grobs, grobs being graphical objects.

### `h_data_plot`

```{r, h_data_plot}
#' Helper function: tidy survival fit
#'
#' Convert the survival fit data into a dataframe designed for plotting.
#'
h_data_plot <- function(fit_km,
                        xticks = NULL) {
  y <- broom::tidy(fit_km)
  y$censor <- ifelse(y$n.censor > 0, y$estimate, NA)
  if (!is.null(xticks)) {
    y <- subset(y, time <= max(xticks))
  }
  if (is.null(fit_km$strata)) {
    y$strata <- "Data"
  }
  y
}
```

```{r h_data_plot_example}
library(random.cdisc.data)
library(dplyr)
library(survival)
radtte(cached = TRUE) %>%
  filter(PARAMCD == "OS") %>%
  survfit(form = Surv(AVAL, 1 - CNSR) ~ ARMCD, data = .) %>%
  h_data_plot()
```

### `h_xticks`

```{r h_xticks}
#' Helper function: x tick positions
#'
#' Calculate the positions of ticks on the x-axis. However, if `xticks` already
#' exists it is kept as is. It is based on the same function `ggplot2` relies on,
#' and is required in the graphic and the patient-at-risk annotation table.
#'
h_xticks <- function(data, xticks = NULL) {
  if (is.null(xticks)) {
    labeling::extended(range(data$time)[1], range(data$time)[2], m = 5)
  } else {
    xticks
  }
}
```

```{r h_xticks_examples}
library(random.cdisc.data)
library(dplyr)
library(survival)
data <- radtte(cached = TRUE) %>%
  filter(PARAMCD == "OS") %>%
  survfit(form = Surv(AVAL, 1 - CNSR) ~ ARMCD, data = .) %>%
  h_data_plot()

h_xticks(data)
```


### `h_ggkm`


```{r, h_ggkm}
#' Helper function: KM plot
#'
#' Draw the Kaplan-Meier plot using `ggplot2`.
#'
h_ggkm <- function(data,
                   xticks = NULL,
                   censor_show,
                   xlab,
                   ylab,
                   title) {
  gg <- {
    ggplot2::ggplot(
      data = data,
      mapping = ggplot2::aes(
        x = time,
        y = estimate,
        ymin = conf.low,
        ymax = conf.high,
        color = strata,
        fill = strata
      )
    ) + ggplot2::geom_hline(
      yintercept = 0
    ) + ggplot2::geom_ribbon(
      alpha = .3, lty = 0
    ) + ggplot2::geom_step(
    ) + ggplot2::coord_cartesian(
      ylim = c(0, 1)
    ) + ggplot2::labs(
      x = xlab, y = ylab, title = title
    )
  }
  if (censor_show) {
    gg <- gg +
      ggplot2::geom_point(
        data = subset(data, !is.na(censor)),
        ggplot2::aes(x = time, y = censor),
        shape = "|"
      )
  }
  if (!is.null(xticks)) {
    gg <- gg + ggplot2::scale_x_continuous(breaks = xticks)
  }
  gg + ggplot2::theme_minimal(
  ) + ggplot2::theme(
    legend.position = "bottom",
    legend.title = ggplot2::element_blank(),
    panel.grid.major.x = ggplot2::element_line(size = 2)
  )
}
```


```{r, h_ggkm_example}
library(random.cdisc.data)
library(dplyr)
library(survival)
fit_km <- radtte(cached = TRUE) %>%
  filter(PARAMCD == "OS") %>%
  survfit(form = Surv(AVAL, 1 - CNSR) ~ ARMCD, data = .)
data_plot <- h_data_plot(fit_km = fit_km)
xticks <- h_xticks(data = data_plot)
gg <- h_ggkm(
  data = data_plot,
  censor_show = TRUE,
  xticks = xticks, xlab = "Days", ylab = "Survival Probability",
  title = "tt"
)

gg
```



### `h_decompose_gg`

```{r, h_decompose_gg}
#' `ggplot` Decomposition
#'
#' The elements composing the `ggplot` are extracted and organised in a
#' list containing:
#' the panel (`panel`),
#' the y-axis and its label (`yaxis`, `ylab`),
#' idem for the x-axis (`xaxis`, `xlab`),
#' the legend (`guide`).
#'

h_decompose_gg <- function(gg) {
  g_el <- ggplot2::ggplotGrob(gg)
  y <- c(
    panel = "panel", yaxis = "axis-l", xaxis = "axis-b",
    xlab = "xlab-b", ylab = "ylab-l", guide = "guide"
  )
  lapply(X = y, function(x) gtable::gtable_filter(g_el, x))
}

```

```{r, h_decompose_gg_example}
library(random.cdisc.data)
library(dplyr)
library(survival)
library(grid)

fit_km <- radtte(cached = TRUE) %>%
  filter(PARAMCD == "OS") %>%
  survfit(form = Surv(AVAL, 1 - CNSR) ~ ARMCD, data = .)
data_plot <- h_data_plot(fit_km = fit_km)
xticks <- h_xticks(data = data_plot)
gg <- h_ggkm(
  data = data_plot,
  censor_show = TRUE,
  xticks = xticks, xlab = "Days", ylab = "Survival Probability",
  title = "tt"
)

g_el <- h_decompose_gg(gg)
grid.newpage()
grid.rect(gp = gpar(lty = 1, col = "red", fill = "gray85", lwd = 5))
grid.draw(g_el$panel)

grid.newpage()
grid.rect(gp = gpar(lty = 1, col = "royalblue", fill = "gray85", lwd = 5))
grid.draw(with(g_el, cbind(ylab, yaxis)))

```

Layout
------

With the data and after decomposition of the ggplot into elements, it is
then possible to calculate the layout that will frame the final KM plot.

### `h_km_layout`

The layout corresponds to a grid of two columns and five rows of unequal
dimensions.

- The left column gets the annotation of the ggplot (y-axis) and the
names of the strata for the patient at risk tabulation.
The main constraint is about the width of the columns which must allow the
writing of the strata name.
- The right column receive the ggplot, the legend, the x-axis and the
patient at risk table.

Most of the dimension are fixed, only the curve is flexible and will accommodate
with the remaining free space.

```{r, h_km_layout}
#' Helper: KM Layout
#'
#' Prepares a (5 rows) x (2 cols) layout for the Kaplan-Meier curve.
#'
h_km_layout <- function(data_plot, g_el) {
  txtlines <- levels(as.factor(data_plot$strata))
  nlines <- nlevels(as.factor(data_plot$strata))
  col_annot_width <- max(
    c(
      as.numeric(grid::convertX(g_el$yaxis$width + g_el$ylab$width, "pt")),
      as.numeric(
        grid::convertX(
          grid::stringWidth(txtlines) + grid::unit(7, "pt"), "pt"
        )
      )
    )
  )
  grid::grid.layout(
    nrow = 5, ncol = 2,
    widths = grid::unit(c(col_annot_width, 1), c("pt", "null")),
    heights = grid::unit(
      c(
        1,
        grid::convertX(with(g_el, xaxis$height + ylab$width), "pt"),
        grid::convertX(g_el$guide$heights, "pt"),
        nlines + 1,
        grid::convertX(with(g_el, xaxis$height + ylab$width), "pt")
      ),
      c(
        "null",
        "pt",
        "pt",
        "lines",
        "pt"
      )
    )
  )
}
```

```{r, h_km_layout_example}
fit_km <- radtte(cached = TRUE) %>%
  filter(PARAMCD == "OS") %>%
  survfit(form = Surv(AVAL, 1 - CNSR) ~ ARMCD, data = .)
data_plot <- h_data_plot(fit_km = fit_km)
xticks <- h_xticks(data = data_plot)
gg <- h_ggkm(
  data = data_plot,
  censor_show = TRUE,
  xticks = xticks, xlab = "Days", ylab = "Survival Probability",
  title = "tt"
)
g_el <- h_decompose_gg(gg)
lyt <- h_km_layout(data_plot = data_plot, g_el = g_el)
grid.show.layout(lyt)
```


Patient at risk
---------------

A difficulty is to plan the table which will be aligned below the KM curve.

### `h_grob_tbl_at_risk`

```{r, h_grob_tbl_at_risk}
#' Helper: Patient-at-Risk Grobs
#'
#' Two grobs are obtained, one corresponding to row labelling and the
#' second to the number of patient at risk
#'
h_grob_tbl_at_risk <- function(data_plot, annot_tbl) {
  txtlines <- levels(as.factor(data_plot$strata))
  nlines <- nlevels(as.factor(data_plot$strata))
  vp_table <- grid::plotViewport(margins = grid::unit(c(0, 0, 0, 0), "lines"))
  gb_table_left_annot <- grid::gList(
    grid::rectGrob(
      x = 0, y = grid::unit(c(1:nlines) - 1, "lines"),
      gp = grid::gpar(fill = c("gray95", "gray90"), alpha = 1, col = "white"),
      height = grid::unit(1, "lines"), just = "bottom", hjust = 0
    ),
    grid::textGrob(
      label = unique(annot_tbl$strata),
      x = .95,
      y = grid::unit(unique(as.numeric(annot_tbl$strata)) - .5, "native"),
      hjust = 1,
      gp = grid::gpar(fontface = "italic", fontsize = 10)
    )
  )
  gb_patient_at_risk <- grid::gList(
    grid::rectGrob(
      x = 0, y = grid::unit(c(1:nlines) - 1, "lines"),
      gp = grid::gpar(fill = c("gray95", "gray90"), alpha = 1, col = "white"),
      height = grid::unit(1, "lines"), just = "bottom", hjust = 0
    ),
    grid::textGrob(
      label = annot_tbl$n.risk,
      x = grid::unit(annot_tbl$time, "native"),
      y = grid::unit(as.numeric(annot_tbl$strata) - .5, "line") # maybe native
    )
  )
  list(
    at_risk = grid::gList(
      grid::gTree(
        vp = vp_table,
        children = grid::gList(
          grid::gTree(
            vp = grid::dataViewport(
              xData = data_plot$time,
              yscale = c(0, nlines + 1),
              extension = c(0.05, 0)
            ),
            children = grid::gList(gb_patient_at_risk)
          )
        )
      )
    ),
    label = grid::gList(
      grid::gTree(
        vp = grid::viewport(width = max(grid::stringWidth(txtlines))),
        children = grid::gList(
          grid::gTree(
            vp = grid::dataViewport(
              xscale = 0:1,
              yscale = c(0, nlines + 1),
              extension = c(0.0, 0)
            ),
            children = grid::gList(gb_table_left_annot)
          )
        )
      )
    )
  )
}
```


```{r, h_grob_tbl_at_risk_example}
fit_km <- radtte(cached = TRUE) %>%
  filter(PARAMCD == "OS") %>%
  survfit(form = Surv(AVAL, 1 - CNSR) ~ ARMCD, data = .)

data_plot <- h_data_plot(fit_km = fit_km)

xticks <- h_xticks(data = data_plot)

gg <- h_ggkm(
  data = data_plot,
  censor_show = TRUE,
  xticks = xticks, xlab = "Days", ylab = "Survival Probability",
  title = "tt"
)

# The annotation table reports the patient at risk for a given strata and
# time (`xticks`).
annot_tbl <- summary(fit_km, time = xticks)
annot_tbl <- if (is.null(fit_km$strata)) {
  with(annot_tbl, data.frame(n.risk = n.risk, time = time, strata = "Data"))
} else {
  with(annot_tbl, data.frame(n.risk = n.risk, time = time, strata = strata))
}

# The annotation table is transformed into a grob.
tbl <- h_grob_tbl_at_risk(data_plot = data_plot, annot_tbl = annot_tbl)

# For the representation, the layout is estimated for which the decomposition
# of the graphic element is necessary.
g_el <- h_decompose_gg(gg)
lyt <- h_km_layout(data_plot = data_plot, g_el = g_el)

grid.newpage()
pushViewport(viewport(layout = lyt, height = .95, width = .95))
grid.rect(gp = gpar(lty = 1, col = "purple", fill = "gray85", lwd = 1))
pushViewport(viewport(layout.pos.row = 4, layout.pos.col = 2))
grid.rect(gp = gpar(lty = 1, col = "orange", fill = "gray85", lwd = 1))
grid.draw(tbl$at_risk)
popViewport()
pushViewport(viewport(layout.pos.row = 4, layout.pos.col = 1))
grid.rect(gp = gpar(lty = 1, col = "green3", fill = "gray85", lwd = 1))
grid.draw(tbl$label)
```

## Median of Survival

### `h_tbl_median_surv`

```{r, h_tbl_median_surv}
#' Helper Function: Survival Estimations
#'
#' Transform a survival fit to a table with groups in rows characterised
#' by N, median and 95% confidence interval.
#'
h_tbl_median_surv <- function(fit) {
  y <- if (is.null(fit$strata)) {
    as.data.frame(t(summary(fit)$table), row.names = "Data")
  } else {
    as.data.frame(summary(fit)$table)
  }
  y$records <- round(y$records)
  y$median <- signif(y$median, 4)
  y$`CI` <- paste0( # nolint
    "(", signif(y$`0.95LCL`, 4), ", ", signif(y$`0.95UCL`, 4), ")"
  )

  setNames(
    y[c("records", "median", "CI")],
    c("N", "Median", "95% CI")
  )
}

```

```{r, h_tbl_median_surv_example}
library(random.cdisc.data)
library(dplyr)

adtte <- radtte(cached = TRUE) %>%
  filter(PARAMCD == "OS")

fit <- survival::survfit(
  form = survival::Surv(AVAL, 1 - CNSR) ~ ARMCD,
  data = adtte
)
h_tbl_median_surv(fit = fit)
```

### `h_grob_median_surv`

```{r, h_grob_median_surv}
#' Helper Function: Survival Estimation Grob
#'
#' The survival fit is transformed in a grob containing a table with groups in
#' rows characterised by N, median and 95% confidence interval.
#'
h_grob_median_surv <- function(fit,
                               ttheme = gridExtra::ttheme_default()) {
  data <- h_tbl_median_surv(fit) # nolint
  gt <- gridExtra::tableGrob(d = data, theme = ttheme)
  vp <- grid::viewport(
    x = grid::unit(1, "npc"),
    y = grid::unit(1, "npc"),
    height = sum(gt$heights),
    width = sum(gt$widths),
    just = c("right", "top")
  )

  grid::gList(
    grid::gTree(
      vp = vp,
      children = grid::gList(gt)
    )
  )
}

```

```{r, h_grob_median_surv_example}
library(random.cdisc.data)
library(dplyr)
library(survival)

grid.newpage()
grid.rect(gp = gpar(lty = 1, col = "pink", fill = "gray85", lwd = 1))
radtte(cached = TRUE) %>%
  filter(PARAMCD == "OS") %>%
  survfit(form = Surv(AVAL, 1 - CNSR) ~ ARMCD, data = .) %>%
  h_grob_median_surv %>%
  grid.draw
```


## Y-Axis Annotation

### `h_grob_y_annot`

```{r, h_grob_y_annot}
#' Helper: Grid Object with y-axis Annotation
#'
#' Build the y-axis annotation from a decomposed `ggplot`.
#'
h_grob_y_annot <- function(ylab, yaxis) {
  grid::gList(
    grid::gTree(
      vp = grid::viewport(
        width = grid::convertX(yaxis$width + ylab$width, "pt"),
        x = grid::unit(1, "npc"),
        just = "right"
      ),
      children = grid::gList(cbind(ylab, yaxis))
    )
  )
}

```


```{r, h_grob_y_annot_example}
library(random.cdisc.data)
library(dplyr)

fit_km <- radtte(cached = TRUE) %>%
  filter(PARAMCD == "OS") %>%
  survfit(form = Surv(AVAL, 1 - CNSR) ~ ARMCD, data = .)
data_plot <- h_data_plot(fit_km = fit_km)
xticks <- h_xticks(data = data_plot)
gg <- h_ggkm(
  data = data_plot,
  censor_show = TRUE,
  xticks = xticks, xlab = "Days", ylab = "Survival Probability",
  title = "tt"
)

g_el <- h_decompose_gg(gg)

grid.newpage()
pvp <- plotViewport(margins = c(5, 4, 2, 20))
pushViewport(pvp)
grid.draw(h_grob_y_annot(ylab = g_el$ylab, yaxis = g_el$yaxis))
grid.rect(gp = gpar(lty = 1, col = "gray35", fill = NA))
```

## Final Assembly

### `g_km`

```{r, g_km}
g_km <- function(fit_km,
                 xticks = NULL,
                 # col = NA,
                 # lty = 1,
                 # lwd = 1,
                 censor_show = TRUE,
                 # pch = 3,
                 # size = grid::unit(0.5, "char"),
                 # max_time = NULL,
                 xlab = "Days",
                 ylab = "Survival Probability",
                 title = "Kaplan-Meier Plot",
                 draw = TRUE,
                 newpage = TRUE
                 # gp = NULL,
                 # vp = NULL,
                 # name = NULL,
) {
  data_plot <- h_data_plot(fit_km = fit_km, xticks = xticks) # nolint
  xticks <- h_xticks(data = data_plot, xticks = xticks) # nolint
  gg <- h_ggkm(
    data = data_plot,
    censor_show = censor_show,
    xticks = xticks,
    xlab = xlab,
    ylab = ylab,
    title = title
  )
  g_el <- h_decompose_gg(gg) # nolint
  # This is the content of the table that will be below the graph.
  annot_tbl <- summary(fit_km, time = xticks)
  annot_tbl <- if (is.null(fit_km$strata)) {
    data.frame(
      n.risk = annot_tbl$n.risk,
      time = annot_tbl$time,
      strata = "Data"
    )
  } else {
    data.frame(
      n.risk = annot_tbl$n.risk,
      time = annot_tbl$time,
      strata = annot_tbl$strata
    )
  }

  grobs_patient <- h_grob_tbl_at_risk(
    data_plot = data_plot, annot_tbl = annot_tbl
  )

  lyt <- h_km_layout(data_plot = data_plot, g_el = g_el)

  result <- grid::gTree(
    vp = grid::viewport(layout = lyt, height = .95, width = .95),
    children = grid::gList(

      # The Kaplan - Meier curve (top-right corner).
      grid::gTree(
        vp = grid::viewport(layout.pos.row = 1, layout.pos.col = 2),
        children = grid::gList(g_el$panel)
      ),

      # Survfit summary table (top-right corner).
      grid::gTree(
        vp = grid::viewport(layout.pos.row = 1, layout.pos.col = 2),
        children = h_grob_median_surv(fit = fit_km)
      ),

      # Add the y-axis annotation (top-left corner).
      grid::gTree(
        vp = grid::viewport(layout.pos.row = 1, layout.pos.col = 1),
        children = h_grob_y_annot(ylab = g_el$ylab, yaxis = g_el$yaxis)
      ),

      # Add the x-axis annotation (second row below the Kaplan Meier Curve).
      grid::gTree(
        vp = grid::viewport(layout.pos.row = 2, layout.pos.col = 2),
        children = grid::gList(rbind(g_el$xaxis, g_el$xlab))
      ),

      # Add the legend.
      grid::gTree(
        vp = grid::viewport(layout.pos.row = 3, layout.pos.col = 2),
        children = grid::gList(g_el$guide)
      ),

      # Add the table with patient-at-risk numbers.
      grid::gTree(
        vp = grid::viewport(layout.pos.row = 4, layout.pos.col = 2),
        children = grobs_patient$at_risk
      ),

      # Add the table with patient-at-risk labels.
      grid::gTree(
        vp = grid::viewport(layout.pos.row = 4, layout.pos.col = 1),
        children = grobs_patient$label
      ),

      # Add the x-axis for the table.
      grid::gTree(
        vp = grid::viewport(layout.pos.row = 5, layout.pos.col = 2),
        children = grid::gList(rbind(g_el$xaxis, g_el$xlab))
      )
    )
  )

  if (newpage & draw)  grid::grid.newpage()
  if (draw) grid::grid.draw(result)

  return(
    list(
      lyt = lyt,
      result = result
    )
  )
}

```

### Examples {.tabset}

#### Default

```{r, g_km_example_1}
fit_km <- radtte(cached = TRUE) %>%
  filter(PARAMCD == "OS") %>%
  survfit(form = Surv(AVAL, 1 - CNSR) ~ ARMCD, data = .)
res <- g_km(fit_km = fit_km)
```

#### Layout

```{r, g_km_example_2}
grid.show.layout(res$lyt)
```

#### `xticks`

```{r, g_km_example_3, fig.width = 8, fig.height = 9}
res <- g_km(fit_km = fit_km, xticks = seq(0, 2000, 365))
```


```{r}
get_simple <- function(n = 200) {
  set.seed(n, kind = "Mersenne-Twister")
  data.frame(
    time = sample(seq(0, 10, 2), size = 2 * n, replace = TRUE),
    status = sample(x = c(0, 1), 2 * n, replace = TRUE),
    armcd  = factor(
      LETTERS[
        rep(c(1, 2), each = n)
        ], levels = c("A", "B")
      )
  )
}

fit_km <- survfit(Surv(time, status) ~ armcd, data = get_simple())
res <- g_km(fit_km = fit_km, xticks =  seq(0, 10, 2))
h_data_plot(fit_km)
```

