---
title: "Allow MMRM with a single treatment arm"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data

```{r}
library(assertthat)
library(dplyr)
library(random.cdisc.data)

packageVersion("tern")
packageVersion("rtables")

adsl <- radsl(cached = TRUE)
adqs <- radqs(cached = TRUE)
adqs_f <- adqs %>%
  dplyr::filter(PARAMCD == "FKSI-FWB" & !AVISIT %in% c("BASELINE")) %>%
  droplevels() %>%
  dplyr::mutate(ARMCD = factor(ARMCD, levels = c("ARM B", "ARM A", "ARM C"))) %>%
  dplyr::mutate(
    AVISITN = rank(AVISITN) %>%
    as.factor() %>%
    as.numeric() %>%
    as.factor()
  )
var_labels(adqs_f) <- var_labels(adqs)

adqs_single <- adqs_f %>%
  dplyr::filter(ARMCD == "ARM B") %>%
  dplyr::mutate(ARMCD = droplevels(ARMCD))
adsl_single <- adsl %>%
  dplyr::filter(ARMCD == "ARM B")
```

## Model fitting

Let's see what we need to change to make the model fitting work. Basically we want to treat the case of no arm variable explicitly.

### Check function

Now only when the arm variable is specified, we check it.

```{r}
check_mmrm_vars <- function(vars,
                            data) {
  stopifnot(is.list(vars))
  stopifnot(is.data.frame(data))
  labels <- list()

  # Check whether variables are specified and in data.
  is_specified <- function(var) {
    !is.null(vars[[var]])
  }
  is_specified_and_in_data <- function(var) {
    is_specified(var) && all(vars[[var]] %in% names(data))
  }
  check_and_get_label <- function(var) {
    stopifnot(is_specified_and_in_data(var))
    res <- NULL
    for (v in vars[[var]]) {
      label <- attr(data[[v]], "label")
      string <- ifelse(!is.null(label), label, v)
      res <- c(res, setNames(string, v))
    }
    labels[[var]] <<- res  # Saves the result under the `var` element of `labels`.
  }
  check_and_get_label("response")
  check_and_get_label("id")
  if (is_specified("arm")) {
    check_and_get_label("arm")
  }
  check_and_get_label("visit")
  if (is_specified("covariates")) {
    vars$parts <- unique(unlist(strsplit(vars$covariates, split = "\\*|:")))
    check_and_get_label("parts")
  }

  # Subset data to observations with complete regressors.
  regressor_vars <- c(vars$arm, vars$visit, vars$parts)
  data_complete_regressors <- data %>%
    dplyr::filter(complete.cases(data[, regressor_vars])) %>%
    droplevels()

  # Check variable values in this complete data set.
  response_values <- data_complete_regressors[[vars$response]]
  stopifnot(is.numeric(response_values))

  if (is_specified("arm")) {
    arm_values <- data_complete_regressors[[vars$arm]]
    assertthat::assert_that(is_df_with_nlevels_factor(
      data_complete_regressors,
      variable = vars$arm,
      n_levels = 2,
      relation = ">="
    ))
  }

  visit_values <- data_complete_regressors[[vars$visit]]
  stopifnot(is.factor(visit_values))

  # Remove all entries where response is NA, droplevels as well.
  data_complete <- data_complete_regressors %>%
    filter(!is.na(data_complete_regressors[, vars$response])) %>%
    droplevels()

  if (is_specified("arm")) {
    # Check all arms will still be present after NA filtering.
    assertthat::assert_that(is_df_with_nlevels_factor(
      data_complete,
      variable = vars$arm,
      n_levels = nlevels(arm_values),
      relation = "=="
    ))

    # Each arm should have at least have 5 records.
    if (!all(table(data_complete[[vars$arm]]) > 5)) {
      stop(paste("Each group / arm should have at least 5 records with non-missing", vars$response))
    }
  } else {
    # The data should have at least 5 records.
    if (!(nrow(data_complete) > 5)) {
      stop(paste("There should be at least 5 records with non-missing", vars$response))
    }
  }

  return(labels)
}
```

### Build formula

Here in case of missing arm variable we let the arm/visit part only contain the visit variable.

```{r}
build_mmrm_formula <- function(
  vars,
  cor_struct = c(
    "unstructured",
    "random-quadratic",
    "random-slope",
    "compound-symmetry"
  )
) {
  cor_struct <- match.arg(cor_struct)

  covariates_part <- paste(
    vars$covariates,
    collapse = " + "
  )

  arm_visit_part <- if (is.null(vars$arm)) {
    vars$visit
  } else {
    paste0(
      vars$arm,
      "*",
      vars$visit
    )
  }

  random_effects_part <- cor_struct %>%
    switch(
      "unstructured" = "(0 + visit_var | id_var)",
      "random-quadratic" = "(poly(as.numeric(visit_var), df=2) | id_var)",
      "random-slope" = "(as.numeric(visit_var) | id_var)",
      "compound-symmetry" = "(1 | id_var)"
    ) %>%
    gsub("visit_var", vars$visit, x = .) %>% #nolint
    gsub("id_var", vars$id, x = .) #nolint

  rhs_formula <- paste(
    arm_visit_part,
    "+",
    random_effects_part
  )
  if (covariates_part != "") {
    rhs_formula <- paste(
      covariates_part,
      "+",
      rhs_formula
    )
  }

  full_formula <- as.formula(paste(
    vars$response,
    "~",
    rhs_formula
  ))

  return(full_formula)
}
```

### Extract LS means

We need to adjust the formula specifying the LS means when there is no arm variable. Further, we only calculate contrasts when there is an arm variable.

```{r}
get_mmrm_lsmeans <- function(fit,
                             vars,
                             conf_level,
                             weights) {
  data_complete <- fit@frame
  specs <- if (is.null(vars$arm)) {
    as.formula(paste("~ 1 |", vars$visit))
  } else {
    as.formula(paste("~ ", vars$arm, "|", vars$visit))
  }

  lsmeans <- emmeans(
    fit,
    mode = "satterthwaite",
    specs = specs,
    weights = weights,
    data = data_complete,
    # The below option is needed to enable analysis of more than 3000 observations.
    lmerTest.limit = nrow(data_complete)
  )

  # Relative Reduction (in change from baseline) is calculated using model based
  # LS means as 100*(LS mean change from baseline in Control Pooled group –
  # LS mean change from baseline in Treatment Group)/LS mean change from
  # baseline in Control Pooled group.

  # Adjusted estimate for each arm.
  estimates <- confint(lsmeans, level = conf_level) %>%
    as.data.frame() %>%
    rename(
      estimate = .data$emmean,
      se = .data$SE,
      lower_cl = .data$lower.CL,
      upper_cl = .data$upper.CL
    )

  data_n <- data_complete %>%
    group_by_at(.vars = c(vars$visit, vars$arm)) %>%
    summarise(n = n()) %>%
    ungroup()

  estimates <- suppressWarnings(
    # We don't have labels in `estimates`, which triggers a warning.
    estimates %>%
      left_join(data_n, by = c(vars$visit, vars$arm))
  )

  results <-  if (is.null(vars$arm)) {
    list(
      estimates = estimates
    )
  } else {
    # Get LS means for reference group to join into full dataframe so that relative reduction in
    # LS mean (mean of response variable) can be computed with respect to reference level (e.g. ARM A).
    arm_levels <- levels(data_complete[[vars$arm]])
    reference_level <- arm_levels[1]
    means_at_ref <- estimates %>%
      filter(!!as.symbol(vars$arm) == reference_level) %>%
      select(c(vars$visit, "estimate")) %>%
      rename(ref = .data$estimate)

    relative_reduc <- estimates %>%
      filter(!!as.symbol(vars$arm) != reference_level) %>%
      left_join(means_at_ref, by = c(vars$visit)) %>%
      mutate(relative_reduc = (.data$ref - .data$estimate) / .data$ref) %>%
      select(c(vars$visit, vars$arm, "relative_reduc"))

    # Start with the differences between LS means.
    sum_fit_diff <- summary(
      contrast(lsmeans, method = "trt.vs.ctrl", parens = NULL),
      level = conf_level,
      infer = c(TRUE, TRUE),
      adjust = "none"
    )

    # Get the comparison group name from "contrast" column,
    # e.g. "ARMB - ARMA" shall return "ARMB", i.e. remove " - ARMA" part.
    contrasts <- sum_fit_diff %>%
      mutate(
        col_by = factor(
          gsub(paste0("\\s-\\s", reference_level), "", .data$contrast),
          levels = arm_levels
        )
      ) %>%
      select(-.data$contrast) %>%
      rename(!!as.symbol(vars$arm) := .data$col_by) %>%
      left_join(relative_reduc, by = c(vars$visit, vars$arm)) %>%
      mutate(!!as.symbol(vars$arm) := droplevels(!!as.symbol(vars$arm), exclude = reference_level)) %>%
      rename(
        se = .data$SE,
        lower_cl = .data$lower.CL,
        upper_cl = .data$upper.CL,
        t_stat = .data$t.ratio,
        p_value = .data$p.value
      )

    list(
      estimates = estimates,
      contrasts = contrasts
    )
  }

  return(results)
}
```

### Fitting function

Now for the fitting function we just have to be careful for the `ref_level` result list element.

```{r}
fit_mmrm <- function(
  vars = list(
    response = "AVAL",
    covariates = c(),
    id = "USUBJID",
    arm = "ARM",
    visit = "AVISIT"
  ),
  data,
  conf_level = 0.95,
  cor_struct = "unstructured",
  weights_emmeans = "proportional",
  optimizer = "automatic",
  parallel = FALSE
) {
  labels <- check_mmrm_vars(vars, data)
  assertthat::assert_that(
    is_proportion(conf_level),
    assertthat::is.flag(parallel)
  )
  formula <- build_mmrm_formula(vars, cor_struct)

  fit <- fit_lme4(
    formula = formula,
    data = data,
    optimizer = optimizer,
    n_cores = ifelse(parallel, get_free_cores(), 1L)
  )

  lsmeans <- get_mmrm_lsmeans(
    fit = fit,
    vars = vars,
    conf_level = conf_level,
    weights = weights_emmeans
  )

  cov_estimate <- get_lme4_cov_estimate(fit)
  id_rows <- which(fit@frame[[vars$id]] == attr(cov_estimate, "id"))
  visits <- fit@frame[[vars$visit]][id_rows]
  rownames(cov_estimate) <- colnames(cov_estimate) <- as.character(visits)

  diagnostics <- get_lme4_diagnostics(fit, cov_est = cov_estimate)

  results <- list(
    fit = fit,
    cov_estimate = cov_estimate,
    diagnostics = diagnostics,
    lsmeans = lsmeans,
    vars = vars,
    labels = labels,
    ref_level = if (is.null(vars$arm)) NULL else levels(data[[vars$arm]])[1],
    conf_level = conf_level
  )
  class(results) <- "mmrm"
  return(results)
}
```

## Test

Now we can try it out:

```{r}
mmrm_results <- fit_mmrm(
  vars = list(
    response = "AVAL",
    covariates = "BMRKR2",
    id = "USUBJID",
    visit = "AVISIT"
  ),
  data = adqs_single,
  cor_struct = "unstructured",
  weights_emmeans = "equal"
)
```

So we have our `mmrm` object here.

## Tables

The fixed effect estimates are fine:

```{r}
as.rtable(mmrm_results, type = "fixed")
```

Covariance too:

```{r}
as.rtable(mmrm_results, type = "cov")
```

Criteria too:

```{r}
as.rtable(mmrm_results, type = "diagnostic")
```

### LS means

The LS means table don't work out of the box.

#### tidy method

First we have to adapt the tidy method.

```{r}
tidy.mmrm <- function(x) {  #nolint
  vars <- x$vars
  estimates <- x$lsmeans$estimates
  df <- if (is.null(vars$arm)) {
    nams <- names(estimates)
    to_rename <- match(c("estimate", "se", "df", "lower_cl", "upper_cl"), nams)
    names(estimates)[to_rename] <- paste0(names(estimates)[to_rename], "_est")
    estimates
  } else {
    contrasts <- x$lsmeans$contrasts
    contrasts[[vars$arm]] <- factor(contrasts[[vars$arm]], levels = levels(estimates[[vars$arm]]))
    merge(
      x = estimates,
      y = contrasts,
      by = c(vars$arm, vars$visit),
      suffixes = c("_est", "_contr"),
      all = TRUE
    )
  }
  df$conf_level <- x$conf_level
  as_tibble(df)
}
```

#### Stats function

We need to define new a stats function and the corresponding formatted analysis function because the argument `.in_ref_col` in the original one.

```{r}
s_mmrm_lsmeans_single <- function(df) {
  list(
    n = df$n,
    adj_mean_se = c(df$estimate_est, df$se_est),
    adj_mean_ci = with_label(c(df$lower_cl_est, df$upper_cl_est), f_conf_level(df$conf_level))
  )
}

a_mmrm_lsmeans_single <- make_afun(
  s_mmrm_lsmeans_single,
  .labels = c(
    adj_mean_se = "Adjusted Mean (SE)"
  ),
  .formats = c(
    n = "xx.",
    adj_mean_se = sprintf_format("%.3f (%.3f)"),
    adj_mean_ci = "(xx.xxx, xx.xxx)"
  ),
  .indent_mods = c(
    adj_mean_ci = 1L
  )
)
```

#### Layout function

Here we now decide between using the existing or the new formatted analysis function.
```{r}
summarize_lsmeans <- function(lyt,
                              arms = TRUE,
                              ...,
                              table_names = "lsmeans_summary",
                              .stats = NULL,
                              .formats = NULL,
                              .indent_mods = NULL,
                              .labels = NULL) {
  afun <- make_afun(
    ifelse(arms, a_mmrm_lsmeans, a_mmrm_lsmeans_single),
    .stats = .stats,
    .formats = .formats,
    .indent_mods = .indent_mods,
    .labels = .labels
  )
  analyze(
    lyt = lyt,
    vars = "n",
    afun = afun,
    table_names = table_names,
    extra_args = list(...)
  )
}
```

#### Try out

We can then use the new tidy method and the updated layout function. Note that we can also have the estimates as columns here.

```{r}
df <- tidy(mmrm_results)

basic_table() %>%
  split_rows_by("AVISIT") %>%
  add_colcounts() %>%
  summarize_lsmeans(arms = FALSE) %>%
  build_table(
    df = df,
    alt_counts_df = adsl_single
  )

basic_table() %>%
  split_cols_by("AVISIT") %>%
  summarize_lsmeans(arms = FALSE) %>%
  build_table(
    df = df
  )
```


## Graphs

This will work similarly. Can work on the details in production.
