---
title: "Design for helper function to create layout map"
author: "Rosemary Li"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: journal
    highlight: zenburn
vignette: >
  %\VignetteIndexEntry{Design of LBT04}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "svg"
)
```

Configuration
=============

```{r setup}
library(rtables)
library(tern)
library(dplyr)
library(assertthat)

packageVersion("tern")
packageVersion("rtables")
```

Introduction
=============

When templates like LBT04, VST02, EGT02, LBT15 et al were refactored using `trim_levels_to_map`, how to define the map
correctly is critical. Different templates have different scenarios to create the map. For example
- VST02/ECGT02 have both directions
- LBT04 follow the rule that [at least one observation with ANRLO > 0 for low direction and at least one observation with ANRHI is not missing for high direction](https://pages.github.roche.com/MDIS/stream_doc/2_11/um/macro_ref/MACRO_str_tlg_proc_createanrindref.html).
- LBT15 map is created from metadata, a similar rule can be found at https://pages.github.roche.com/MDIS/stream_doc/um/techdoc/macro_ref/MACRO_str_tlg_proc_addgradedirs.html?highlight=addncigradedirs

Design
=============

Maybe the function can get different `method` so that map for different templates can be created accordingly.

