---
title: "Multi-variable Logistic Regression with Interaction (LGRT02)"
author: "Heng Wang, DSB"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-variable Logistic Regression with Interaction (LGRT02)}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "svg"
)
```

```{r data, results = "asis", eval = TRUE}
library(tern)
library(random.cdisc.data)
library(dplyr)
library(rtables)
library(assertthat)

ADRS <- radrs(cached = TRUE) #nolint
adrs_labels <- formatable::var_labels(ADRS) #nolint

ADRS_f <- ADRS %>% #nolint
  dplyr::filter(PARAMCD == "BESRSPI") %>%
  dplyr::filter(RACE %in% c("ASIAN", "WHITE", "BLACK OR AFRICAN AMERICAN")) %>%
  dplyr::mutate(
    Response = case_when(AVALC %in% c("PR", "CR") ~ 1, TRUE ~ 0),
    RACE = factor(RACE),
    SEX = factor(SEX)
  )

formatable::var_labels(ADRS_f) <- c(adrs_labels, "Response" = "Tumor Response")
```

Now let's have a look at the interaction table. Basically we need another hierarchy level here in the tidied dataset and the row splits, let's call it "reference" since we are referring to a level or value of the interaction term (either treatment arm or the covariate which is interacting with treatment arm). Then the content rows at the "term" level no longer display odds ratio and CI for that, since that is then shown at the "reference" level.

Plus we have at the bottom of the interaction table a section where the interaction terms themselves are tested. This is similar to the other "term" level summaries in this table in that no odds ratio or CI are displayed. The values there can be obtained from the usual `summary` of the `glm` fit, since they refer to single coefficient estimates only.

Finally we still have covariates that are not interacting with treatment arm where the same tabular summary as without interactions at all in the model is shown.

# Current baseline

Let's see what we currently have in `tern` already.

We can fit and extract the main and interaction term estimates:
```{r}
model_factor_glm <- fit_logistic(
  data = ADRS_f,
  variables = list(
   response = "Response",
   arm = "ARMCD",
   covariates = c("AGE", "SEX"),
   interaction = "SEX"
  )
)
glm_simple_term_extract("ARMCD", model_factor_glm)
glm_simple_term_extract("SEX", model_factor_glm)
glm_interaction_extract("ARMCD:SEX", model_factor_glm)

model_numeric_glm <- fit_logistic(
  data = ADRS_f,
  variables = list(
   response = "Response",
   arm = "ARMCD",
   covariates = c("AGE", "SEX"),
   interaction = "AGE"
  )
)
glm_simple_term_extract("ARMCD", model_numeric_glm)
glm_simple_term_extract("AGE", model_numeric_glm)
glm_interaction_extract("ARMCD:AGE", model_numeric_glm)
```

That means we have all the numbers for the columns "Degrees of Freedom", "Parameter Estimate", "Standard Error", and "p-value". We are only missing the odds ratio estimates and corresponding CIs.

# Odds ratio estimation

The main challenge her is that we also need to account for the covariance between the coefficients. Therefore the calculations are not as simple as for the non-interaction case, and we therefore don't include them in the helper functions directly but factor them out in dedicated functions.

## Calculating the estimates

### Categorical covariate

Let's start simple with a categorical covariate interacting with the treatment arm.

```{r}
h_get_interaction_vars <- function(fit_glm) {  # done
  assertthat::assert_that("glm" %in% class(fit_glm))
  terms_name <- attr(terms(fit_glm), "term.labels")
  xs_class <- attr(fit_glm$terms, "dataClasses")
  interaction_term <- terms_name[which(!terms_name %in% names(xs_class))]
  strsplit(interaction_term, split = ":")[[1]]
}

h_interaction_coef_name <- function( # done
  interaction_vars,
  first_var_with_level,
  second_var_with_level
) {
  first_name <- paste(first_var_with_level, collapse = "")
  second_name <- paste(second_var_with_level, collapse = "")
  if (first_var_with_level[1] == interaction_vars[1]) {
    paste(first_name, second_name, sep = ":")
  } else if (second_var_with_level[1] == interaction_vars[1]) {
    paste(second_name, first_name, sep = ":")
  } else {
    stop("something is wrong")
  }
}

h_odds_ratio_categorical_interaction <- function( #done, renamed to h_or_cat_interaction
  odds_ratio_var,  # name of the variable for which to calculate odds ratios
  interaction_var,   # name of the interaction variable
  fit_glm,
  conf_level = 0.95) {
  interaction_vars <- h_get_interaction_vars(fit_glm)
  assertthat::assert_that(
    assertthat::is.string(odds_ratio_var),
    assertthat::is.string(interaction_var),
    all(c(odds_ratio_var, interaction_var) %in% interaction_vars),
    identical(length(interaction_vars), 2L)
  )
  xs_level <- fit_glm$xlevels
  xs_coef <- coef(fit_glm)
  xs_vcov <- vcov(fit_glm)
  y <- list()
  for (var_level in xs_level[[odds_ratio_var]][-1]) {
    x <- list()
    for (ref_level in xs_level[[interaction_var]]) {
      coef_names <- paste0(odds_ratio_var, var_level)
      if (ref_level != xs_level[[interaction_var]][1]) {
        interaction_coef_name <- h_interaction_coef_name(
          interaction_vars,
          c(odds_ratio_var, var_level),
          c(interaction_var, ref_level)
        )
        coef_names <- c(
          coef_names,
          interaction_coef_name
        )
      }
      if (length(coef_names) > 1) {
        ones <- t(c(1, 1))
        est <- as.numeric(ones %*% xs_coef[coef_names])
        se <- sqrt(as.numeric(ones %*% xs_vcov[coef_names, coef_names] %*% t(ones)))
      } else {
        est <- xs_coef[coef_names]
        se <- sqrt(as.numeric(xs_vcov[coef_names, coef_names]))
      }
      or <- exp(est)
      ci <- exp(est + c(-1, 1) * qnorm((1 + conf_level) / 2) * se)
      x[[ref_level]] <- list(or = or, ci = ci)
    }
    y[[var_level]] <- x
  }
  y
}
```

So for example:
```{r}
h_odds_ratio_categorical_interaction(odds_ratio_var = "ARMCD", interaction_var = "SEX", model_factor_glm)
h_odds_ratio_categorical_interaction(odds_ratio_var = "SEX", interaction_var = "ARMCD", model_factor_glm)
```

### Continuous covariate

Now we look at a continuous covariate. Here we need an additional argument to specify increments that is either `NULL` in which case we would just look at the median of the interaction variable.

```{r}
h_odds_ratio_continuous_interaction <- function( #done, renamed to h_or_cont_interaction
  odds_ratio_var,  # name of the variable for which to calculate odds ratios
  interaction_var,   # name of the interaction variable
  fit_glm,
  interaction_increments = NULL, # optional increments for interaction variable
  conf_level = 0.95) {
  interaction_vars <- h_get_interaction_vars(fit_glm)
  assertthat::assert_that(
    assertthat::is.string(odds_ratio_var),
    assertthat::is.string(interaction_var),
    all(c(odds_ratio_var, interaction_var) %in% interaction_vars),
    identical(length(interaction_vars), 2L)
  )
  
  checkmate::assert_numeric(interaction_increments, min.len = 1, null.ok = TRUE, any.missing = FALSE)

  xs_level <- fit_glm$xlevels
  xs_coef <- coef(fit_glm)
  xs_vcov <- vcov(fit_glm)
  xs_class <- attr(fit_glm$terms, "dataClasses")
  model_data <- fit_glm$model
  if (!is.null(interaction_increments)){
    assertthat::assert_that(
      xs_class[interaction_var] == "numeric"
    )
  }
  y <- list()
  if (xs_class[interaction_var] == "numeric") {
    if (is.null(interaction_increments)) {
      interaction_increments <- ceiling(median(model_data[[interaction_var]]))
    }

    for (var_level in xs_level[[odds_ratio_var]][-1]) {
      x <- list()
      for (increment in interaction_increments) {
        coef_names <- paste0(odds_ratio_var, var_level)
        if (increment != 0) {
          interaction_coef_name <- h_interaction_coef_name(
            interaction_vars,
            c(odds_ratio_var, var_level),
            c(interaction_var, "")
          )
          coef_names <- c(
            coef_names,
            interaction_coef_name
          )
        }
        if (length(coef_names) > 1) {
          xvec <- t(c(1, increment))
          est <- as.numeric(xvec %*% xs_coef[coef_names])
          se <- sqrt(as.numeric(xvec %*% xs_vcov[coef_names, coef_names] %*% t(xvec)))
        } else {
          est <- xs_coef[coef_names]
          se <- sqrt(as.numeric(xs_vcov[coef_names, coef_names]))
        }
        or <- exp(est)
        ci <- exp(est + c(-1, 1) * qnorm((1 + conf_level) / 2) * se)
        x[[as.character(increment)]] <- list(or = or, ci = ci)
      }
      y[[var_level]] <- x
    }
  } else {
    assertthat::assert_that(
      xs_class[odds_ratio_var] == "numeric",
      xs_class[interaction_var] == "factor"
    )
    for (var_level in xs_level[[interaction_var]]) {
      coef_names <- odds_ratio_var
      if (var_level != xs_level[[interaction_var]][1]) {
        interaction_coef_name <- h_interaction_coef_name(
          interaction_vars,
          c(odds_ratio_var, ""),
          c(interaction_var, var_level)
        )
        coef_names <- c(
          coef_names,
          interaction_coef_name
        )
      }
      if (length(coef_names) > 1) {
        xvec <- t(c(1, 1))
        est <- as.numeric(xvec %*% xs_coef[coef_names])
        se <- sqrt(as.numeric(xvec %*% xs_vcov[coef_names, coef_names] %*% t(xvec)))
      } else {
        est <- xs_coef[coef_names]
        se <- sqrt(as.numeric(xs_vcov[coef_names, coef_names]))
      }
      or <- exp(est)
      ci <- exp(est + c(-1, 1) * qnorm((1 + conf_level) / 2) * se)
      y[[var_level]] <- list(or = or, ci = ci)
    }
  }
  y
}
```

Try it out:

```{r}
h_odds_ratio_continuous_interaction(odds_ratio_var = "ARMCD", interaction_var = "AGE",
                                    interaction_increments = c(20, 30, 40), conf_level = 0.95,
                                    fit_glm = model_numeric_glm)
```


### Wrapper that checks for class

```{r}
h_odds_ratio_interaction <- function( #done, renamed to h_or_interaction
  odds_ratio_var,  # name of the variable for which to calculate odds ratios
  interaction_var,   # name of the interaction variable
  fit_glm,
  conf_level = 0.95,
  interaction_increments = NULL) {
  xs_class <- attr(fit_glm$terms, "dataClasses")
  ret <- if (any(xs_class[c(odds_ratio_var, interaction_var)] == "numeric")) {
    h_odds_ratio_continuous_interaction(
      odds_ratio_var, interaction_var, fit_glm,
      conf_level = conf_level, interaction_increments = interaction_increments
    )
  } else if (all(xs_class[c(odds_ratio_var, interaction_var)] == "factor")) {
    h_odds_ratio_categorical_interaction(odds_ratio_var, interaction_var, fit_glm, conf_level = conf_level)
  } else {
    stop("wrong interaction variable class")
  }
  return(ret)
}
```

So for example:
```{r}
h_odds_ratio_interaction(odds_ratio_var = "ARMCD", interaction_var = "AGE", model_numeric_glm,
                         interaction_increments = c(20, 30, 40))

h_odds_ratio_interaction(odds_ratio_var = "AGE", interaction_var = "ARMCD", model_numeric_glm)
```

There is a lot of duplicate code in the above functions, which we can further refactor during production.

# Tidying the estimates into data frame

Now we need to tackle the formatting of the estimates in a logistic regression model with interaction terms into a data frame.

## Baseline

Currently we have only the simple terms version:

```{r}
glm_simple_term_extract("ARMCD", model_factor_glm)
glm_interaction_extract("ARMCD:SEX", model_factor_glm)
h_logistic_simple_terms("AGE", model_factor_glm)
```

We don't yet have something like the `h_logistic_simple_terms` which works on terms that are involved in interactions.
As mentioned above, in our helper function here we will need to produce additional columns `interaction` and `reference` that mention the name and the reference value of the interaction variable.

## simple term extract

We also need to adapt `glm_simple_term_extract` to return the right columns `is_variable_summary` and `is_term_summary`.

```{r}
glm_simple_term_extract <- function(x, fit_glm) { #done, renamed to h_glm_simple_term_extract
  assertthat::assert_that(
    "glm" %in% class(fit_glm),
    assertthat::is.string(x)
  )
  xs_class <- attr(fit_glm$terms, "dataClasses")
  xs_level <- fit_glm$xlevels
  xs_coef <- summary(fit_glm)$coefficients
  stats <- c("estimate" = "Estimate", "std_error" = "Std. Error", "pvalue" = "Pr(>|z|)")
  # Make sure x is not interaction term
  assertthat::assert_that(x %in% names(xs_class))
  x_sel <- if (xs_class[x] == "numeric") x else paste0(x, xs_level[[x]][-1])
  x_stats <- as.data.frame(xs_coef[x_sel, stats, drop = FALSE], stringsAsFactors = FALSE)
  colnames(x_stats) <- names(stats)
  x_stats$estimate <- as.list(x_stats$estimate)
  x_stats$std_error <- as.list(x_stats$std_error)
  x_stats$pvalue <- as.list(x_stats$pvalue)
  x_stats$df <- as.list(1)
  if (xs_class[x] == "numeric") {
    x_stats$term <- if (xs_class[x] == "numeric") x else xs_level[[x]][-1]
    x_stats$is_variable_summary <- TRUE
    x_stats$is_term_summary <- FALSE
  } else {
    x_stats$term <- xs_level[[x]][-1]
    x_stats$is_variable_summary <- FALSE
    x_stats$is_term_summary <- TRUE
    main_effects <- car::Anova(fit_glm, type = 3, test.statistic = "Wald")
    x_main <- data.frame(
      pvalue = main_effects[x, "Pr(>Chisq)", drop = TRUE],
      term = paste(x, "Reference =", xs_level[[x]][1]),
      df = main_effects[x, "Df", drop = TRUE],
      stringsAsFactors = FALSE
    )
    x_main$pvalue <- as.list(x_main$pvalue)
    x_main$df <- as.list(x_main$df)
    x_main$estimate <- list(numeric(0))
    x_main$std_error <- list(numeric(0))
    if (length(xs_level[[x]][-1]) == 1) {
      x_main$pvalue <- list(numeric(0))
      x_main$df <- list(numeric(0))
    }
    x_main$is_variable_summary <- TRUE
    x_main$is_term_summary <- FALSE
    x_stats <- rbind(x_main, x_stats)
  }
  x_stats$variable <- x
  rownames(x_stats) <- NULL
  x_stats[c(
    "variable",
    "term",
    "estimate",
    "std_error",
    "df",
    "pvalue",
    "is_variable_summary",
    "is_term_summary"
  )]
}
```

Try it out:

```{r}
glm_simple_term_extract("SEX", model_numeric_glm)
glm_simple_term_extract("AGE", model_factor_glm)
```

## interaction extract

Similarly we also need to adapt the interaction extraction function to get the right flag columns in.

```{r}
glm_interaction_extract <- function(x, fit_glm) { #done, renamed to h_glm_interaction_extract
  assertthat::assert_that("glm" %in% class(fit_glm))
  terms_name <- attr(terms(fit_glm), "term.labels")
  xs_class <- attr(fit_glm$terms, "dataClasses")
  assertthat::assert_that(
    assertthat::is.string(x),
    # Only deal with one interaction term
    identical(x, terms_name[which(!terms_name %in% names(xs_class))])
  )
  vars <- unlist(strsplit(x, ":"))
  # Only take two-way interaction
  assertthat::assert_that(
    length(vars) == 2,
    # Only consider simple case: first variable in interaction is arm, a categorical variable
    xs_class[vars[1]] != "numeric"
  )
  xs_level <- fit_glm$xlevels
  xs_coef <- summary(fit_glm)$coefficients
  main_effects <- car::Anova(fit_glm, type = 3, test.statistic = "Wald")
  stats <- c("estimate" = "Estimate", "std_error" = "Std. Error", "pvalue" = "Pr(>|z|)")
  v1_comp <- xs_level[[vars[1]]][-1]
  if (xs_class[vars[2]] == "numeric") {
    x_stats <- as.data.frame(
      xs_coef[paste0(vars[1], v1_comp, ":", vars[2]), stats, drop = FALSE],
      stringsAsFactors = FALSE
    )
    colnames(x_stats) <- names(stats)
    x_stats$term <- v1_comp
  } else if (xs_class[vars[2]] != "numeric") {
    v2_comp <- xs_level[[vars[2]]][-1]
    x_sel <- paste0(levels(interaction(paste0(vars[1], v1_comp), paste0(vars[2], v2_comp), sep = ":")))
    x_stats <- as.data.frame(xs_coef[x_sel, stats, drop = FALSE], stringsAsFactors = FALSE)
    colnames(x_stats) <- names(stats)
    x_stats$term <- paste0(levels(interaction(v1_comp, v2_comp, sep = " * ")))
  }
  x_stats$df <- as.list(1)
  x_stats$pvalue <- as.list(x_stats$pvalue)
  x_stats$is_variable_summary <- FALSE
  x_stats$is_term_summary <- TRUE
  x_main <- data.frame(
    pvalue = main_effects[x, "Pr(>Chisq)", drop = TRUE],
    term = paste(vars[1], vars[2], sep = " * "),
    df = main_effects[x, "Df", drop = TRUE],
    stringsAsFactors = FALSE
  )
  x_main$pvalue <- as.list(x_main$pvalue)
  x_main$df <- as.list(x_main$df)
  x_main$estimate <- list(numeric(0))
  x_main$std_error <- list(numeric(0))
  x_main$is_variable_summary <- TRUE
  x_main$is_term_summary <- FALSE

  x_stats <- rbind(x_main, x_stats)
  x_stats$variable <- x
  rownames(x_stats) <- NULL
  x_stats[c("variable", "term", "estimate", "std_error", "df", "pvalue", "is_variable_summary", "is_term_summary")]
}
```

Try it out:

```{r}
glm_interaction_extract("ARMCD:SEX", model_factor_glm)
glm_interaction_extract("ARMCD:AGE", model_numeric_glm)
```

## interaction term extract

And finally we need something new as well.

```{r}
glm_inter_term_extract <- function(odds_ratio_var, #done, renamed to h_glm_inter_term_extract
                                   interaction_var,
                                   fit_glm,
                                   ...) {
  # first we can get the main effects
  main_stats <- glm_simple_term_extract(odds_ratio_var, fit_glm)
  main_stats$interaction <- NA
  main_stats$reference <- NA
  main_stats$is_reference_summary <- FALSE
  main_stats$odds_ratio <- NA
  main_stats$lcl <- NA
  main_stats$ucl <- NA

  # then we get the odds ratio estimates and put into df form
  or_numbers <- h_odds_ratio_interaction(odds_ratio_var, interaction_var, fit_glm, ...)
  is_num_or_var <- attr(fit_glm$terms, "dataClasses")[odds_ratio_var] == "numeric"

  # this if/else below can be simplified in production
  if (is_num_or_var) {
    # numeric OR variable
    references <- names(or_numbers)
    n_ref <- length(references)

    extract_from_list <- function(l, name, pos = 1) {
      unname(unlist(
        lapply(or_numbers, function(x) {
          x[[name]][pos]
        })
      ))
    }
    or_stats <- data.frame(
      variable = odds_ratio_var,
      term = odds_ratio_var,
      interaction = interaction_var,
      reference = references,
      estimate = NA,
      std_error = NA,
      odds_ratio = extract_from_list(or_numbers, "or"),
      lcl = extract_from_list(or_numbers, "ci", pos = 1),
      ucl = extract_from_list(or_numbers, "ci", pos = 2),
      df = NA,
      pvalue = NA,
      is_variable_summary = FALSE,
      is_term_summary = FALSE,
      is_reference_summary = TRUE
    )
  } else {
    # categorical OR variable
    references <- names(or_numbers[[1]])
    n_ref <- length(references)

    extract_from_list <- function(l, name, pos = 1) {
      unname(unlist(
        lapply(or_numbers, function(x) {
          lapply(x, function(y) y[[name]][pos])
        })
      ))
    }
    or_stats <- data.frame(
      variable = odds_ratio_var,
      term = rep(names(or_numbers), each = n_ref),
      interaction = interaction_var,
      reference = unlist(lapply(or_numbers, names)),
      estimate = NA,
      std_error = NA,
      odds_ratio = extract_from_list(or_numbers, "or"),
      lcl = extract_from_list(or_numbers, "ci", pos = 1),
      ucl = extract_from_list(or_numbers, "ci", pos = 2),
      df = NA,
      pvalue = NA,
      is_variable_summary = FALSE,
      is_term_summary = FALSE,
      is_reference_summary = TRUE
    )
  }

  df <- rbind(
    main_stats[, names(or_stats)],
    or_stats
  )

  df[order(-df$is_variable_summary, df$term, -df$is_term_summary, df$reference), ]
}
```

Let's try it out:

```{r}
glm_inter_term_extract(odds_ratio_var = "ARMCD",
                       interaction_var = "SEX",
                       model_factor_glm,
                       conf_level = 0.95)

glm_inter_term_extract(odds_ratio_var = "ARMCD",
                       interaction_var = "AGE",
                       model_numeric_glm,
                       conf_level = 0.95,
                       interaction_increments = c(20, 30, 40))

glm_inter_term_extract(odds_ratio_var = "AGE",
                       interaction_var = "ARMCD",
                       model_numeric_glm,
                       conf_level = 0.95)
# note: must not use interaction increments here, since that only works the other way around
```


## data frame production

Since in our case we have the odds ratios already, we can directly write the analogue to `h_logistic_simple_terms`.

```{r}
h_logistic_inter_terms <- function(x, fit_glm, conf_level,  #done
                                   interaction_increments = NULL) {
  # find out the interaction variables and interaction term.
  inter_vars <- h_get_interaction_vars(fit_glm)
  assertthat::assert_that(identical(length(inter_vars), 2L))
  inter_term_index <- intersect(grep(inter_vars[1], x), grep(inter_vars[2], x))
  inter_term <- x[inter_term_index]

  # first for the non-interaction vars we need the standard stuff.
  normal_terms <- setdiff(x, union(inter_vars, inter_term))

  # do the normal stuff. -> we need to put this in helper function in production to not
  # duplicate with `h_logistic_simple_terms`
  x_stats <- lapply(normal_terms, glm_simple_term_extract, fit_glm)
  x_stats <- do.call(rbind, x_stats)
  q_norm <- qnorm((1 + conf_level) / 2)
  x_stats$odds_ratio <- lapply(x_stats$estimate, exp)
  x_stats$lcl <- Map(function(or, se) exp(log(or) - q_norm * se), x_stats$odds_ratio, x_stats$std_error)
  x_stats$ucl <- Map(function(or, se) exp(log(or) + q_norm * se), x_stats$odds_ratio, x_stats$std_error)
  normal_stats <- x_stats
  normal_stats$interaction <- NA
  normal_stats$reference <- NA
  normal_stats$is_reference_summary <- FALSE

  # now the interaction term itself.
  inter_term_stats <- glm_interaction_extract(inter_term, fit_glm)
  inter_term_stats$odds_ratio <- NA
  inter_term_stats$lcl <- NA
  inter_term_stats$ucl <- NA
  inter_term_stats$interaction <- NA
  inter_term_stats$reference <- NA
  inter_term_stats$is_reference_summary <- FALSE

  is_intervar1_numeric <- attr(fit_glm$terms, "dataClasses")[inter_vars[1]] == "numeric"

  # now proceed with the interaction stuff.
  inter_stats_one <- glm_inter_term_extract(
    inter_vars[1],
    inter_vars[2],
    fit_glm,
    conf_level = conf_level,
    interaction_increments = `if`(is_intervar1_numeric, NULL, interaction_increments)
  )
  inter_stats_two <- glm_inter_term_extract(
    inter_vars[2],
    inter_vars[1],
    fit_glm,
    conf_level = conf_level,
    interaction_increments = `if`(is_intervar1_numeric, interaction_increments, NULL)
  )

  # put everything together
  col_names <- c(
    "variable",
    "term",
    "interaction",
    "reference",
    "estimate",
    "std_error",
    "df",
    "pvalue",
    "odds_ratio",
    "lcl",
    "ucl",
    "is_variable_summary",
    "is_term_summary",
    "is_reference_summary"
  )
  df <- rbind(
    normal_stats[, col_names],
    inter_stats_one[, col_names],
    inter_stats_two[, col_names],
    inter_term_stats[, col_names]
  )
  df$ci <- combine_vectors(df$lcl, df$ucl)
  df
}
```

Let's try it out:

```{r}
# factor interaction
h_logistic_inter_terms(
  x = attr(terms(model_factor_glm), "term.labels"),
  fit_glm = model_factor_glm,
  conf_level = 0.95
)

# continuous interaction
h_logistic_inter_terms(
  x = attr(terms(model_numeric_glm), "term.labels"),
  fit_glm = model_numeric_glm,
  conf_level = 0.95,
  interaction_increments = c(20, 30)
)
```


## Tidy method

We need to adapt `tidy.glm` to use the new helper function:

```{r, tidy_glm}
tidy.glm <- function(fit_glm, conf_level = 0.95, ...) { #nolint #done
  assertthat::assert_that(
    "glm" %in% class(fit_glm),
    fit_glm$family$family == "binomial"
  )
  terms_name <- attr(terms(fit_glm), "term.labels")
  xs_class <- attr(fit_glm$terms, "dataClasses")
  interaction <- terms_name[which(!terms_name %in% names(xs_class))]
  if (length(interaction) == 0) {
    h_logistic_simple_terms(x = terms_name, fit_glm = fit_glm, conf_level = conf_level)
  } else {
    h_logistic_inter_terms(x = terms_name, fit_glm = fit_glm, conf_level = conf_level, ...)
  }
}
```

Let's try it out:

```{r try_tidy}
df <- broom::tidy(model_factor_glm)
df

df_num <- broom::tidy(model_numeric_glm, interaction_increments = c(20, 30))
df_num
```


# Tabulation

## Stats function

See below, we will just use content functions that format the right data frame entries.

## Layout creating functions and table code

In the end we should be able to run something like this:

```{r, eval = FALSE}
adrs <- get_adrs()
mod1 <- fit_logistic(
  adrs,
  variables = list(
    response = "Response",
    arm = "ARMCD",
    covariates = c("SEX", "RACE", "AGE"),
    interaction = "SEX"
  )
)
df <- broom::tidy(mod1)
result <- basic_table() %>%
  split_rows_by("variable") %>%
  split_rows_by("term", split_fun = drop_split_levels) %>%
  split_rows_by("interaction", split_fun = drop_split_levels) %>%
  split_rows_by("reference", split_fun = drop_split_levels) %>%
  summarize_logistic(conf_level = 0.95) %>%
  build_table(df = df)
```

Alternatively we could try something like this, which seems more manageable as we can use the `summarize_rows_groups` for each split to display the information:

```{r, eval = FALSE}
result <- basic_table() %>%
  logistic_cols(...) %>%
  # this sets up the split_cols_by_multivar structure with all variables.
  # note that we don't need flexibility here regarding which columns to have
  split_rows_by("variable", labels_var = "variable_label") %>%
  # since the labels are currently not yet as we need them with `n`, parentheses etc.
  summarize_logistic_variable_test() %>%
  # this would use `summarize_row_groups` internally. so the user can choose if they want this or not
  split_rows_by("term", labels_var = "term_label", split_fun = drop_split_levels) %>%
  summarize_logistic_term_estimates() %>%
  # same like above
  # then if there is interaction in the model/df can proceed with:
  split_rows_by("interaction", labels_var = "interaction_label", split_fun = drop_split_levels) %>%
  split_rows_by("reference", labels_var = "reference_label", split_fun = drop_split_levels) %>%
  summarize_logistic_odds_ratios() %>%
  build_table(df = df)
```

Let's try this quickly:

Here we have a layout creating function that just sets up the multiple columns.
Note that we need to supply the confidence level for the CI column here, which is not optimal but ok.

```{r logistic_cols}
logistic_cols <- function(lyt, conf_level = 0.95) { # done, renamed to logistic_regression_cols
  vars <- c("df", "estimate", "std_error", "odds_ratio", "ci", "pvalue")
  var_labels <- c(
    df = "Degrees of Freedom", estimate = "Parameter Estimate",
    std_error = "Standard Error", odds_ratio = "Odds Ratio",
    ci = paste("Wald", tern:::f_conf_level(conf_level)), pvalue = "p-value"
  )
  split_cols_by_multivar(
    lyt = lyt,
    vars = vars,
    varlabels = var_labels
  )
}
```

Now here we create 3 different layout functions which are only different in which flag column they use to identify the row they should use to summarize the split.

```{r summarize_logistic_funs}
# Helper function to give an empty cell in case only NA in this entry.
# need the `unlist(x)` for the CI entries.
zeroNA <- function(x) { #nolint # done, renamed to unlist_and_blank_na
  `if`(all(is.na(unlist(x))), character(), unlist(x))
}

# content function which just gives df$var at the row identified by
# flag variable in the given format.
ident_cfun <- function(var, flag, format = "xx") { # done, renamed to cfun_by_flag
  force(var)
  force(format)
  function(df, labelstr, .df_row) {
    row_index <- which(.df_row[[flag]])
    formatable::with_label(
      rcell(zeroNA(df[[var]][row_index]), format = format),
      labelstr
    )
  }
}

# constructor which produces layout creating function for given flag.
create_summary_layout_fun <- function(flag) {
  force(flag)
  function(lyt) {
    cfun_list <- list(
      df = ident_cfun("df", flag, format = "xx."),
      estimate = ident_cfun("estimate", flag, format = "xx.xx"),
      std_error = ident_cfun("std_error", flag, format = "xx.xx"),
      odds_ratio = ident_cfun("odds_ratio", flag, format = "xx.xx"),
      ci = ident_cfun("ci", flag, format = "(xx.xx, xx.xx)"),
      pvalue = ident_cfun("pvalue", flag, format = "x.xxxx | (<0.0001)")
    )
    summarize_row_groups(
      lyt = lyt,
      cfun = cfun_list
    )
  }
}

# since we have 3 hierarchy levels (variable, term, reference) we need 3 layout creating functions.
summarize_logistic_variable_test <- create_summary_layout_fun("is_variable_summary") #nolint
summarize_logistic_term_estimates <- create_summary_layout_fun("is_term_summary") #nolint
summarize_logistic_odds_ratios <- create_summary_layout_fun("is_reference_summary") #nolint
```

This custom split function constructor combines `drop` and `remove` features. I will do a PR for rtables for this one since
it uses internal rtables functions and could be generally useful.

```{r splitfun}
drop_and_remove_split_levels <- function(excl) {
  force(excl)
  function(df, spl, vals = NULL, labels = NULL, trim = FALSE) {
    var <- rtables:::spl_payload(spl)
    df2 <- df[!(df[[var]] %in% excl), ]
    df2[[var]] <- factor(df2[[var]])
    rtables:::.apply_split_inner(spl, df2, vals = vals, labels = labels,
                                 trim = trim)
  }
}
```

So all together:

```{r}
df <- broom::tidy(model_factor_glm)

result <- basic_table() %>%
  logistic_cols() %>%
  split_rows_by("variable", split_fun = drop_and_remove_split_levels("")) %>%
  summarize_logistic_variable_test() %>%
  split_rows_by("term", split_fun = drop_and_remove_split_levels("")) %>%
  summarize_logistic_term_estimates() %>%
  split_rows_by("interaction", split_fun = drop_and_remove_split_levels("")) %>%
  split_rows_by("reference", split_fun = drop_and_remove_split_levels("")) %>%
  summarize_logistic_odds_ratios() %>%
  build_table(df)
result
```

So the problem was that the summarize row groups function gets the whole `x` vector from the rows. But we just need the first one / unique one. We need other columns that flag which row is a variable / term / reference summary.

This is what we want. We can further add label columns to the input data frame to polish the presentation.
Now that we know which input data frame we need we can work towards that from the data side above.

We could later wrap this and the other layout functions in another large one which handles all to shorten the default call for the user.

Now for the numeric interaction:

```{r}
df_num <- broom::tidy(model_numeric_glm, interaction_increments = c(20, 30, 40))

result_num <- basic_table() %>%
  logistic_cols() %>%
  split_rows_by("variable", split_fun = drop_and_remove_split_levels("")) %>%
  summarize_logistic_variable_test() %>%
  split_rows_by("term", split_fun = drop_and_remove_split_levels("")) %>%
  summarize_logistic_term_estimates() %>%
  split_rows_by("interaction", split_fun = drop_and_remove_split_levels("")) %>%
  split_rows_by("reference", split_fun = drop_and_remove_split_levels("")) %>%
  summarize_logistic_odds_ratios() %>%
  build_table(df_num)
result_num
```
