---
title: "Design for helper function to create layout map in LBT13/14"
author: "Rosemary Li"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: journal
    highlight: zenburn
vignette: >
  %\VignetteIndexEntry{Design of LBT04/VST02/EGT02/LBT15}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "svg"
)
```

Configuration
=============

```{r setup}
library(rtables)
library(tern)
library(dplyr)
library(assertthat)
library(checkmate)

packageVersion("tern")
packageVersion("rtables")
```

Introduction
=============

When templates like LBT13 and 14 were refactored using `trim_levels_to_map`, we would like to have some helper function
for users to easily create the correct map. The metadata used to create the map can be found at
metadata/__atoxgr_allhigh_ref.sas7bdat

Design
=============

New function to store meta data.
```{r eval=FALSE}
d_atoxgr_allhigh_ref <- function() {
  atoxgr_allhigh_ref <- as.data.frame(
    matrix(c(
      "ALKPHSI",  "1", "ALKPHSI",  "2", "ALKPHSI",  "3", "ALKPHSI",  "4", "ALKPHSI",  "Not High",
      "ALTSI",    "1", "ALTSI",    "2", "ALTSI",    "3", "ALTSI",    "4", "ALTSI",    "Not High",
      "AMYLASSI",    "1", "AMYLASSI",    "2", "AMYLASSI",    "3", "AMYLASSI",    "4", "AMYLASSI",    "Not High",
      "ASTSI",    "1",  "ASTSI",    "2", "ASTSI",    "3", "ASTSI",    "4", "ASTSI",    "Not High",
      "CALCUMSI",    "1", "CALCUMSI",    "2", "CALCUMSI",    "3", "CALCUMSI",    "4", "CALCUMSI",    "Not High",
      "CH0022SI",    "1", "CH0022SI",    "2", "CH0022SI",    "3", "CH0022SI",    "4", "CH0022SI",    "Not High",
      "CH0350SI",    "1", "CH0350SI",    "2", "CH0350SI",    "3", "CH0350SI",    "4", "CH0350SI",    "Not High",
      "CH0365SI",    "1", "CH0365SI",    "2", "CH0365SI",    "3", "CH0365SI",    "4", "CH0365SI",    "Not High",
      "CH0467SI",    "1", "CH0467SI",    "2", "CH0467SI",    "3", "CH0467SI",    "4", "CH0467SI",    "Not High",
      "CHOLESSI",    "1", "CHOLESSI",    "2", "CHOLESSI",    "3", "CHOLESSI",    "4", "CHOLESSI",    "Not High",
      "CPKSI",    "1", "CPKSI",    "2", "CPKSI",    "3",  "CPKSI",    "4",  "CPKSI",    "Not High",
      "CREATNSI",    "1", "CREATNSI",    "2", "CREATNSI",    "3", "CREATNSI",    "4", "CREATNSI",    "Not High",
      "FASTGLSI",    "1", "FASTGLSI",    "2", "FASTGLSI",    "3", "FASTGLSI",    "4", "FASTGLSI",    "Not High",
      "GGTSI",    "1",
      "GGTSI",    "2",
      "GGTSI",    "3",
      "GGTSI",    "4",
      "GGTSI",    "Not High",
      "GLUCSI",    "3",
      "GLUCSI",    "4",
      "GLUCSI",    "Not High",
      "HGBSI",    "1",
      "HGBSI",    "2",
      "HGBSI",    "3",
      "HGBSI",    "Not High",
      "LIPASESI",    "1",
      "LIPASESI",    "2",
      "LIPASESI",    "3",
      "LIPASESI",    "4",
      "LIPASESI",    "Not High",
      "LYMPHSI",    "2",
      "LYMPHSI",    "3",
      "LYMPHSI",    "Not High",
      "MAGNESSI",    "1",
      "MAGNESSI",    "3",
      "MAGNESSI",    "4",
      "MAGNESSI",    "Not High",
      "POTASSI" ,    "1",
      "POTASSI" ,    "2",
      "POTASSI" ,    "3",
      "POTASSI" ,    "4",
      "POTASSI" ,    "Not High",
      "PTINRSI",    "1",
      "PTINRSI",    "2",
      "PTINRSI",    "3",
      "PTINRSI",    "Not High",
      "PTTSI",    "1",
      "PTTSI",    "2",
      "PTTSI",    "3",
      "PTTSI",    "Not High",
      "SODIUMSI",    "1",
      "SODIUMSI",    "2",
      "SODIUMSI",    "3",
      "SODIUMSI",    "4",
      "SODIUMSI",    "Not High",
      "TBILISI",    "1",
      "TBILISI",    "2",
      "TBILISI",    "3",
      "TBILISI",    "4",
      "TBILISI",    "Not High",
      "TRIGSI",    "1",
      "TRIGSI",    "2",
      "TRIGSI",    "3",
      "TRIGSI",    "4",
      "TRIGSI",    "Not High",
      "URACIDSI",    "3",
      "URACIDSI",    "4",
      "URACIDSI",    "Not High",
      "WBCSI",    "3",
      "WBCSI",    "Not High"
    ),
    ncol = 2,
    byrow = TRUE)
  )
  colnames(atoxgr_allhigh_ref) <- c("PARAMCD", "ATOXGR")
  atoxgr_allhigh_ref
}
```

New helper function to create grouping and map? Or it's easier to separate grouping and map?
```{r}
h_toxicity_grade_group <- function(
    df, 
    variables = c("ATOXGR", "BTOXGR"), 
    grouping = list("Not High" = c(0, -1, -2, -3, -4), "1" = 1, "2" = 2, "3" = 3, "4" = 4, "Missing" = "<Missing>")) {
  for (variable in variables) {
    new_var <- paste0(variable, "_GP")
    df[[new_var]] <- NA
    for (i in names(grouping)) {
      df[[new_var]] <- ifelse(df[[variable]] %in% unname(grouping[i])[[1]], i, df[[new_var]])
    }
  }
  df
}

h_toxicity_grade_map <- function(
    paramcds = c("ALKPHSI", "GLUCSI"), 
    ref = d_atoxgr_allhigh_ref(), 
    missing_level = "Missing",
    var_group = c("BTOXGR_GP", "ATOXGR_GP")) {
  ref_sub <- ref %>% filter(PARAMCD %in% paramcds)
  result_df <- data.frame()
  for (paramcd in paramcds) {
    a <- paramcd
    b <- ref_sub %>% filter(PARAMCD == paramcd) %>% pull("ATOXGR")
    b <- c(b, missing_level)
    c <- b
    paramcd_df <- expand.grid(a,b,c)
    colnames(paramcd_df) <- c("PARAMCD", var_group)
    result_df <- rbind(result_df, paramcd_df)
  }
  
  result_df
}
```



Unit tests
=============
```{r eval=FALSE}
ATOXGR <- c(0, 2, 3, 4, -3)
BTOXGR <- c(-1, -3, 0, 3 ,2)
df <- data.frame(ATOXGR, BTOXGR)
h_toxicity_grade_group(df)
```

```{r}
h_map_with_count_abnormal(
  df = df,
  split_rows_varibles = "PARAM",
  anl_variable = "ANRIND",
  abnormal = list(low = "LOW", high = "HIGH"),
  method = "both"
)
```
see difference between default and both
```{r}
df <- df %>% mutate(ANRIND = ifelse(PARAM == "ALT", "NORMAL", as.character(ANRIND)))
df$ANRIND <- factor(df$ANRIND, levels = c("LOW", "HIGH", "NORMAL")) #nolint
h_map_for_count_abnormal(
  df = df,
  variables = list(anl = "ANRIND", split_rows = c("PARAM")),
  abnormal = list(low = "LOW", high = "HIGH"),
  method = "default"
)
h_map_with_count_abnormal(
  df = df,
  split_rows_varibles = "PARAM",
  anl_variable = "ANRIND",
  abnormal = list(low = "LOW", high = "HIGH"),
  method = "both"
)
```

STREAM tests with LBT15
=============

test with new method and stream data for lbt15.
```{r eval=FALSE}
# ADSL and ADLB are stream source data read in with haven::read_sas()

adsl <- ADSL %>% df_explicit_na() %>% filter(SAFFL == "Y")
adlb_f <- ADLB %>% df_explicit_na() %>% filter(ONTRTFL == "Y" & ATOXGR != "<Missing>" & PARCAT2 == "SI") %>%
  mutate(ATOXGR = factor(ATOXGR, levels = c("-4", "4", "-1", "-2", "-3", "0", "1", "2", "3", "<Missing>")),
         BTOXGR = factor(BTOXGR, levels = c("-4", "4", "-1", "-2", "-3", "0", "1", "2", "3", "<Missing>")))
adlb_f <- adlb_f %>%
  mutate(
    ATOXGR_CAT = fct_collapse(ATOXGR,
                              "LOW" = c("-3", "-4"),
                              "NORMAL" = c("-2", "-1", "0", "1", "2"),
                              "HIGH" = c("3", "4")
    ),
    BTOXGR_CAT = fct_collapse(BTOXGR,
                              "LOW" = c("-3", "-4"),
                              "NORMAL" = c("-2", "-1", "0", "1", "2", "<Missing>"),
                              "HIGH" = c("3", "4")
    )
  )


map <- h_map_for_count_abnormal(
  df = adlb_f,
  variables = list(anl = "ATOXGR_CAT", split_rows = c("PARCAT1", "PARAM")),
  abnormal = list(low = c("LOW"), high = c("HIGH")),
  method = "meta_ext_ncictc4",
  na_level = "<Missing>"
)

lyt <- basic_table() %>%
  split_cols_by(var = "ARM") %>%
  add_colcounts() %>%
  split_rows_by("PARCAT1", split_fun = trim_levels_to_map(map = map)) %>%
  split_rows_by("PARAM", split_fun = trim_levels_to_map(map = map)) %>%
  count_abnormal(
    var = "ATOXGR_CAT",
    abnormal = list(low = "LOW", high = "HIGH"),
    variables = list(id = "USUBJID", baseline = "BTOXGR_CAT"),
    exclude_base_abn = TRUE
  ) %>%
  append_topleft("    Direction of Abnormality")

build_table(lyt, adlb_f %>% filter(LBCAT != "<Missing>"), alt_counts_df = adsl)

# variant 2
adlb_f_2 <- adlb_f %>%
  mutate(
    ATOXGR_CAT = fct_collapse(ATOXGR,
                              "LOW" = c("-2", "-3", "-4"),
                              "NORMAL" = c("-1", "0", "1"),
                              "HIGH" = c("2", "3", "4")
    ),
    BTOXGR_CAT = fct_collapse(BTOXGR,
                              "LOW" = c("-2", "-3", "-4"),
                              "NORMAL" = c("-1", "0", "1"),
                              "HIGH" = c("2", "3", "4")
    )
  )

build_table(lyt, adlb_f_2 %>% filter(LBCAT != "<Missing>"), alt_counts_df = adsl)
```

replace stream data with scda so that knit will work
```{r}
ADLB <- haven::read_sas("../adlb.sas7bdat") %>% df_explicit_na() #nolint
adlb_f <- ADLB %>% filter(ONTRTFL == "Y" & ANRIND != "<Missing>" & PARCAT2 == "SI")

h_map_with_count_abnormal(
  df = adlb_f %>% filter(LBCAT != "<Missing>"),
  split_rows_varibles = c("LBCAT", "PARAM"),
  anl_variable = "ANRIND",
  abnormal = list(low = "LOW", high = "HIGH"),
  method = "lbt04"
)

lyt <- basic_table() %>%
  split_cols_by(var = "ARM") %>%
  add_colcounts() %>%
  split_rows_by("PARCAT1", split_fun = trim_levels_to_map(map = map)) %>%
  split_rows_by("PARAM", split_fun = trim_levels_to_map(map = map)) %>%
  count_abnormal(
    var = "ATOXGR_CAT",
    abnormal = list(low = "LOW", high = "HIGH"),
    variables = list(id = "USUBJID", baseline = "BTOXGR_CAT"),
    exclude_base_abn = TRUE
  ) %>%
  append_topleft("    Direction of Abnormality")

build_table(lyt, adlb_f, alt_counts_df = adsl)

```

As discussed in the design PR #255, how should we deal with some PARAMCDs that are not in the input dataset.
```{r}
adlb_f_noALTSI <- adlb_f %>% filter(PARAMCD != "ALTSI")
build_table(lyt, adlb_f_noALTSI %>% filter(LBCAT != "<Missing>"), alt_counts_df = adsl)
```

```{r}
map_test <- h_map_for_count_abnormal(
  df = adlb_f_noALTSI,
  variables = list(anl = "ATOXGR_CAT", split_rows = c("PARCAT1", "PARAM")),
  abnormal = list(low = c("LOW"), high = c("HIGH")),
  method = "meta_ext_ncictc4",
  na_level = "<Missing>"
)

lyt_test <- basic_table() %>%
  split_cols_by(var = "ARM") %>%
  add_colcounts() %>%
  split_rows_by("PARCAT1", split_fun = trim_levels_to_map(map = map_test)) %>%
  split_rows_by("PARAM", split_fun = trim_levels_to_map(map = map_test)) %>%
  count_abnormal(
    var = "ATOXGR_CAT",
    abnormal = list(low = "LOW", high = "HIGH"),
    variables = list(id = "USUBJID", baseline = "BTOXGR_CAT"),
    exclude_base_abn = TRUE
  ) %>%
  append_topleft("    Direction of Abnormality")

build_table(lyt_test, adlb_f_noALTSI %>% filter(LBCAT != "<Missing>"), alt_counts_df = adsl)
```
